Starvation of the bacteria Vibrio atlanticus promotes lightning group-attacks on the dinoflagellate Alexandrium pacificum

Version 1

Abstract

Abstract

Algae serve as a source of nutrients for bacteria in the marine environment. The interactions between algae and bacteria at the phycosphere interface are known to include mutualism, commensalism, competition or antagonism. Here, based on in situ observation and on an in vitro interaction study, we report on a novel form of starvation-induced hunting that the cells of selected Vibrio species exert on the dinoflagellate species. Results showed that Vibrio atlanticus was able to coordinate lightning group attacks then kill the dinoflagellate Alexandrium pacificum ACT03. Briefly, the observed coordinated mechanism of algal-killing consists of first, the ‘immobilization stage’ involving the secretion of algicidal metabolites that disrupt the flagella of the prey. The ‘attack stage’ resembles the ‘wolf-pack attack’ strategy, during which Vibrios surrounds algal cells at high density for a brief period without invading them. Finally, the ‘killing stage’ induces the lysis and degradation of the dinoflagellates. By using a combination of biochemical, proteomic, genetic and fluorescence microscopy approaches, we showed that this relationship is not related to the decomposition of algal organic matter, Vibrio quorum sensing pathways, to the toxicity of the algae or to the pathogenicity of the bacterium but is conditioned by nutrient stress, iron availability and link to the iron-vibrioferrin transport system of V. atlanticus. This is the first evidence of a new mechanism that could be involved in regulating Alexandrium spp. blooms and giving Vibrio a competitive advantage in obtaining nutrients from the environment.

Introduction

Harmful algal blooms (HABs) have experienced an increase in their occurrence, intensity, and geographical distribution on a global scale, resulting in adverse environmental, health, and socioeconomic impacts (Marampouti et al., 2021). HABs have a considerable impact on human health as a result of direct exposure to volatile toxins or by toxic seafood consumption (Burkholder et al., 2018). From an ecological point of view, the expansion of HABs can result in the erosion of biodiversity, because they cause massive mortality of marine species and they are generally monospecific in nature (Chai et al., 2020). In coastal areas, understanding the biological interactions that control toxic algal blooms is therefore a major ecological challenge. Among HABs, a number of Alexandrium species have been placed on the list of invasive Mediterranean species. Among them, Alexandrium pacificum is a flagellated eukaryotic unicellular organism that form with Alexandrium tamarense and Alexandrium fundyense the “Alexandrium tamarense” complex (Hadjadji et al., 2020). Since 1998, A. pacificum (former A. catenella) was monitored by Rephy survey in the Thau lagoon (French Mediterranean) because it produces paralytic shellfish toxins (PSTs) resulting of paralytic shellfish poisoning (PSP) syndrome.

Laania (Laanaia et al., 2013) showed that in Thau lagoon, a water temperature around 20°C for several days and organic and inorganic nutrients in sufficient concentrations are parameters favoring the developemnt of A. pacificum, whose massive blooms occur in autumn.

Algal blooms are seasonal events resulting in a rapid increase in the concentration of a species of algae in an aquatic environment. Depending on the phytoplankton species, the tolerance range of physicochemical parameters is different and influences the time of appearance of blooms (Leblad et al., 2020).

Interestingly, although the collapse of phytoplankton blooms has been previously attributed to viruses (Pal et al., 2020), some ecological studies have suggested an important role of algicidal bacteria (Su et al., 2007; Wang et al., 2010), such as some Vibrio spp. (Li et al., 2014; Wang et al., 2020).

Vibrio (class γ-proteobacteria) are common microorganisms in marine systems worldwide (Baker-Austin et al., 2017; Mavian et al., 2020), where they are important components of the food chain, particularly in biodegradation, nutrient regeneration and biogeochemical cycles (Oberbeckmann et al., 2012). Vibrio is one of the most studied bacterial taxa due to their ubiquity in coastal marine systems and their capacity to cause infections in humans and animals, leading sometimes to epizootic or zoonotic epidemics (LeRoux et al., 2015; Mavian et al., 2020). Vibrio are extremely adaptable to their environment (Johnson, 2013). The main factors influencing their occurrence and distribution in water are temperature, salinity, nutrient availability (Wang et al., 2020), multiple strategies such as biofilm formation on biotic and abiotic surfaces (Espinoza-Vergara et al., 2020), or interactions with a multitude of other organisms such as eukaryotic predators (Drebes Dörr and Blokesch, 2020) or plankton (Lopez-Joven et al., 2018) are used by Vibrio in the environment. There is also evidence that global climate change has increased Vibrio-associated illnesses affecting humans and animals (Brumfield et al., 2021; Muhling et al., 2017). However, the drivers and dynamics of Vibrio survival and propagation in the marine environment are not yet fully understood.

A substantial number of research articles have highlighted the potential of γ-proteobacteria to exert algicidal activity against dinoflagellates, supporting the hypothesis that γ-proteobacteria such as Vibrio play a role in the control of algae blooms in situ (Coyne et al., 2022). However, the mechanisms behind Vibrio-driven algal lysis in the environment remain to be elucidated. Particularly, it is unclear how in the water column, algicidal compounds secreted by bacteria can concentrate around the algae to exert their lytic effect.

Based on observations from the natural environment showing a potent relationship between Vibrio and Alexandrium algae bloom events, this study aim to determine in vitro, the main factors implicated in this relationship. Using a combination of biochemical, proteomic, genetic, and fluorescence microscopy approaches, we explored the role of algal toxicity, bacterial pathogenicity and the quorum sensing pathway on this relationship and showed the important role of nutrient stress and the iron uptake pathway in this unique Vibrio/Alexandrium predator/prey interaction.

Methods

Quantification of Alexandrium algae and Vibrio bacteria in the environment

Seawater samples were collected in the Thau Lagoon (southern France, a shallow Mediterranean ecosystem open to the sea (Abadie et al., n.d.), during spring and autumn 2015. Briefly, samples were collected from the subsurface (-50 cm) near an oyster table at a phytoplankton surveillance site (part of the REPHY network, N 43°26.058’ and E 003°39.878’). Once a week during spring and autumn 2015, during field sampling campaigns, 20 L of water was filtered on board through a 180 μm pore-size nylon membrane. At the laboratory, according to Lopez-Joven et al. (Lopez-Joven et al., 2018) seawater was fractionated into two size classes as follows: 2 L of the above filtrate was filtered through a 0.8 μm pore-size polycarbonate Whatman Nuclepore membrane to obtain organisms in the 0.8–180 μm range corresponding to plankton-associated Vibrio and living Alexandrium forms. Then, the filtrate from the 0.8 µm membrane was filtered again through a 0.2 μm pore-size polycarbonate Whatman Nuclepore membrane until the membrane was saturated. Alexandrium cells, ranging from 25 to 40 µm, belong to the microphytoplankton and are therefore retained in the 0.8–180 µm fraction. Any Vibrio cells potentially associated with or attached to Alexandrium cells will also be retained in this fraction. Vibrio cells are approximately 0.5-0.8 µm-thick. The fraction between 0.2 and 0.8 µm therefore includes the free-living Vibrio. The bacterial population collected on 0.8-µm-pore-size filters was designated the particle-associated community, and the population on 0.2-µm-pore-size filters was designated as the free-living community. Membranes (in triplicate) were then conserved in 500 µL of 100% EtOH at -20°C. Environmental DNA (eDNA) was extracted from the MF Millipore membrane using the Macherey-Nagel NucleoSpin Tissue Kit and resuspended in 100 µL of water. The samples were then stored at -20°C after eDNA quantity and purity were assessed using a NanoDrop system (NanoDrop Technologies, Wilmington, DE, USA). PCR amplification reactions were done on a Roche LightCycler 480 Real-Time thermocycler (qPHD platform, University of Montpellier, France).using specific primer pairs (Table 3). Typically, the reactions contained 1 µL of template DNA (the DNA concentration for all samples varied from 1 to 40 μg mL−1), 0.5 µL of each primer (3.33 µM) and 4 µL of reaction mixture (SYBR Green Master Mix) in a total volume of 6 μL. The reaction parameters were as follows: 5 minutes at 95°C (initial denaturation) and 40 cycles of 10 s at 95°C (denaturation), 10 s at the corresponding hybridization temperature (Table 3) and 10 s at 72°C (elongation). Melting curve profiles were generated by increasing the temperature from 65°C to 95°C at 0.5°C s-1. Amplification products were analysed using LightCycler software (Roche Diagnostics). Vibrio spp. and A. pacificum and A. tamarense were quantified by constructing calibration curves based on DNA from the V. atlanticus LGP32 reference strain (former V. tasmaniensis LGP32) and from the A. pacificum reference strain (ACT03: A. catenella strain isolated form Thau in 2003) and the A. tamarense reference strain (ATT07: A. tamarense isolated from Thau in 2007) (not shown).

Strains and growth conditions

Vibrio strains. Wild-type and isogenic mutants of Vibrio atlanticus LGP32 (Table 1) were used in this study. Deletion-mutants included ΔluxR, ΔluxP, ΔluxM and ΔpvuB isogenic strains. The ΔpvuB mutant was constructed here by allelic exchange as described previously by Le Roux (Le Roux et al., 2007). We also used V. atlanticus LGP32 carrying the pSW3654T-GFP plasmid (Le Roux et al., 2007), hereafter referred to as V. atlanticus LGP32-GFP. Bacterial strains were grown at 22 ± 1°C in Zobell medium (0, 38 µM iron (III)). When needed, 25 μg mL-1 chloramphenicol (Cm) was added to cultures of V. atlanticus LGP32-GFP (Le Roux et al., 2007). Oligonucleotides used for RNA expression analysis are specified in table 3.

Phytoplancton strains. Non-axenic phytoplankton species (Table 2) were grown in batch culture in enriched natural sea water (ENSW, 6,55 µM iron (III)) with a salinity of 36 practical salinity units (PSU) at 22 ± 1°C under cool white fluorescent illumination (100 µmol photons m-2 s-1) and a 12 h:12 h light:dark cycle (Harrison et al., 1980). The algae were used for experiments in their exponential growth phase.

Co-culture assay

For each tested phytoplankton species, 2.104 cells harvested in their exponential growth phase were placed in 20 mL of ENSW medium in a 50 mL suspension culture flask (Cellstar® PS, Greiner bio-one). After incubation for 24 h at 22 ± 1°C under cool white fluorescent illumination, 40 µL of Vibrio strains grown for 12, 36, 60 and 156 h in Zobell medium or in Zobell medium supplemented with Fecl3 (6 µM iron (III)) or with H3BO3 (0.47 mM) or the corresponding culture supernatant were added to the algal culture. After incubation at 22 ± 1°C under cool white fluorescent lights, living, non-swimming, group-attacked and lysed phytoplankton cells were counted in a sedimentation chamber under an inverted microscope. The number of lysed cells corresponded to algae showing disrupted membranes. Non-swimming algae were not counted as lysed cells. For Vibrio analysis, 100 µL of a 1:10 serial dilution mixture in ENSW (from 10-2 to 10-10) was plated on Vibrio Selective TCBS (thiosulfate-citrate-bile salts-sucrose) agar (in triplicate). After incubation for 24 h at 22 ± 1°C, the number of living Vibrio cells was determined by counting colony-forming units (CFUs). Data are the means of three independent experiments.

Microscope observations

The dynamic of the interaction between A. pacificum ACT03 and V. atlanticus LGP32, which is an oyster pathogen isolated from the French Atlantic coast (Gay et al., 2004) and present in the Thau Lagoon, south of France (Lopez-Joven et al., 2018)) was surveyed. As the algal strain used in the study is not axenic, means that additional bacteria, other than the V. atlanticus LGP32, are present in the experiments. To elucidate the interaction without thoroughly accounting for the non-axenic cells, interaction was observed under a Zeiss Axio upright fluorescence microscope equipped with an AxioCamMRm 2 digital microscope camera using V. atlanticus LGP32 tagged with green fluorescent protein (GFP). Lasers were used at excitation wavelength (λex) 488 nm for GFP (emission wavelengths (λem): 505–530 nm) and λex 532 nm for plankton chloroplasts (λem 560–630 nm). Images were taken sequentially to avoid cross-contamination between fluorochromes. Sequences of images were merged during the Vibrio-Alexandrium interaction using ZEN 2012 (blue edition) software. Interaction events between Vibrio strains and phytoplankton strains were also observed under a Leica TCS SPE confocal laser scanning system connected to a Leica DM 2500 upright microscope camera (Montpellier RIO Imaging Platform, University of Montpellier, France).

Comparative proteomic analysis

Vibrio sampling and protein extraction. V. atlanticus LGP32 was grown for 60 h at 22°C in artificial seawater (high nutrient stress, cond. 1) or 12 h at 22°C in Zobell media (low nutrient stress, cond. 2). After 10 min centrifugation at 8000 rpm, crude protein extracts of V. atlanticus LGP32 in each culture condition (triplicates) were obtained by sonication on ice at 20% amplitude for 20 s in 200 µL of ice-cold denaturing buffer (7 M urea, 2 M thiourea, 4% CHAPS in 30 mM Tris-HCl, pH 8.5) and clarified by centrifugation at 2000 x g, 15 min, 4°C. The protein concentration of the supernatant was estimated using the 2D Quant Kit (Cytiva™, MERCK) and samples were stored at -80°C until use.

Bi-dimensional gel electrophoresis (2D gel). Proteins extracts were individually analysed on 2D gel electrophoresis (6 gels per condition each corresponding to different biological replicates). To do so, 100 µg of proteins from each extract was added to rehydration buffer (7 M urea, 2 M thiourea, 4% CHAPS, 65 mM DTT) for a total volume of 350 µL. They were then individually loaded onto 17 cm isoelectric focusing strips (Bio-Rad) with a stabilized non-linear pH ranging from 3 to 10. Due to the high complexity of the protein profile in the acidic part (left) of the gel pH 3–10 (Fig. S3), we conducted additional ‘close-up’ analyses in gels using 17 cm isoelectric focusing strips (Bio-Rad) with a narrower, stabilized pH gradient ranging from 4 to 7. Strips were rehydrated passively for 5 h at 22°C, followed by active rehydration for 14 h under a 50 V current at 22°C (to help large proteins enter the strips). Thereafter, isoelectrofocusing was carried out using the following programme: 50 V for 1 h, 250 V for 1 h, 8000 V for 1 h and a final step at 8000 V for a total of 140 000 V.h with a slow ramping voltage (quadratic increasing voltage) at each step. Focused proteins were reduced by incubating the strip twice in equilibration buffer (1.5 M Tris, 6 M urea, 2% SDS, 30% glycerol; bromophenol blue, pH 8.8) containing DTT (130 mM) at 55°C. Then, they were alkylated by incubation with equilibration buffer containing iodoacetamide (135 mM) on a rocking agitator (400 rpm) at room temperature protected from light. Proteins were also separated according to their molecular weight (second dimension) on 12% acrylamide/0.32% piperazine diacrylamide gels run at 25 mA per gel for 30 min followed by 75 mA per gel for 8 h using a Protean II XL system (Bio-Rad). Gels were stained using an MS-compatible silver staining protocol and scanned using a ChemiDoc MP Imaging System (Bio-Rad) associated with Image Lab software version 4.0.1 (Bio-Rad).

Comparative bioinformatics analysis of 2D gels. Twelve gels (six per condition) were selected for comparative analysis on PD-Quest v. 7.4.0 (Bio-Rad) to identify changes in protein abundance between the proteomic profiles of V. atlanticus LGP32 cultured in contrasting nutrient conditions (ENSW/Zobell). Spots whose mean intensity across six replicates per strain was two times higher or lower than those from the other strain, with a P <0.01 (Mann-Whitney U-test), were considered significantly different in terms of abundance between the two conditions (quantitative difference). Differentially represented spots were then excised from the gels, destained, trypsin-digested and the obtained peptides were identified by tandem mass spectrometry (MS-MS) using the PISSARO platform facility (University of Rouen, France). To identify protein(s) present in each spot, the obtained peptides were compared with V. atlanticus LGP32 reference genome (https://vibrio.biocyc.org/). The genes whose peptides matched strongly were retrieved and used for an BLASTx query against non-redundant databases to determine the protein identity of the best match. A gene was considered as strongly matched when at least two peptides matched the sequence with a coverage of> 6%. Their theoretical isoelectric point (pI) and molecular weights were also calculated using the Expasy server (https://www.expasy.org/) to compare them with the location of the spot on the gel. Altogether, these complementary analyses made it possible to characterize the protein identity of each spot with confidence.

Gene expression analysis

Vibrio sampling and RNA extraction. V. atlanticus LGP32 was grown in Zobell media for 36 h and 60 h at 22°C (decline phase of growth, nutrient stress) or for 12 h at 22°C (exponential grow phase, poor nutrient stress). Total RNA was isolated from V. atlanticus LGP32 using the standard TRIzol method (Invitrogen Life Technologies SAS, Saint-Aubin, France) and then treated with DNase (Invitrogen) to eliminate genomic DNA contamination. After sodium acetate precipitation, the quantity and quality of the total RNA were determined using a NanoDrop spectrophotometer and agarose gel electrophoresis. Following heat denaturing (70°C for 5 minutes), reverse transcription was performed using 1 μg of RNA prepared with 50 ng μL-1 oligo-(dT) 12–18 mer in a 20-μL reaction containing 1 mM dNTPs, 1 unit μL-1 RNAseOUT and 200 units μL-1 Moloney murine leukaemia virus reverse transcriptase (M-MLV RT) in reverse transcriptase buffer, according to the manufacturer’s instructions (Invitrogen Life Technologies SAS, Saint-Aubin, France).

PCR amplification. Amplification reactions were analysed using a Roche LightCycler 480 Real-Time thermocycler (Bio-Environnement platform, University of Perpignan, France). In this study, several PCR primer pairs were designed using Primer3 software (optimal primer size: 20 bases; Tm: 60°C; primer GC%: 50; 2GC clamp and product size range: 150–200 bp) and calibrated with V. atlanticus LGP32 genomic DNA (Table 1). To determine the qPCR efficiency of each primer pair used, standard curves were generated using seven serial dilutions of genomic DNA (10-2, 10-3, 10-4, 10-5, 10-5, 10-7 and 10-8) (not shown); the qPCR efficiencies of the tested genes varied between 1.85 and 2.08 (Table 3). For gene expression, reverse transcription was performed with 1 µg of total RNA using random hexamers and SuperScript IV reverse transcriptase (Invitrogen). The total qPCR reaction volume was 10 μL and consisted of 5 μL of cDNA (diluted 1:5), 2.5 μL of SensiFAST SYBR No-ROX Mix (Bioline) and 100 nM or 300 nM PCR primer pair (Table 3). The reaction parameters were as follows: 2 min at 95°C (initial denaturation) and 40 cycles of 5 s at 95°C (denaturation), 10 s at 59°C (annealing) and 20 s at 72°C (elongation). The specificity of each PCR was checked by measuring fluorescent signals during melting curve analysis (PCR product heated from 65°C to 95°C continuously and slowly at 0.1°C s−1). Relative expression was calculated by normalization to the expression of two constitutively expressed housekeeping genes, namely, 6PKF (VS_2913) and CcmC (VS_0852), using the delta-delta threshold cycle (ΔΔCt) method(Pfaffl, 2001).

Detection of quorum-sensing signalling molecules

Vibrio culture. To detect the QS molecules (AI-2, AI-1 and CAI-1) V. atlanticus LGP32 was grown in Zobell media for 12 h (exponential growth phase, control) and 60 h (decline phase of growth, nutrient stress).

AI-2 analysis. Bioluminescence assay using the QS bioluminescent of Vibrio campbellii MM32 (luxN::Cm, luxS::Tn5Kan) was used to detect AI-2 molecules in culture supernatants. Briefly, Vibrio cultures were centrifuged at 17,000 x g for 10 min, and the resulting supernatants were filtered on 0.22 μm. Then 20 μL of the filtrates were mixed with 180 μL of V. campbellii MM32 diluted 1:5000 then incubated at 30°C and 100 rpm. Luminescence and cell density (OD620) were collected in triplicate and analysed according to Tourneroche et al.(Tourneroche et al., 2019).

AI-1 and CAI-1 extraction and LC-MS analysis. Chemical analyses were conducted with a Q Exactive Focus Orbitrap System coupled to an Ultimate 3000™ ultrahigh-performance liquid chromatography (UHPLC) system (Thermo Fisher Scientific) according to Rodrigues et al. (Rodrigues et al., 2022). Briefly, ethyl acetate (2 mL) was added into each culture (2 mL). This mixture was shaken overnight at room temperature (150 rpm). The two phases were then separated and the aqueous phase was extracted once again. The two obtained organic phases were pooled and the solvent was evaporated under vacuum. The crude extracts were dissolved in 500 µL LC-MS grade methanol for analysis. All experiments were conducted in triplicate. Analyses of extracts and standards (3 μL injected) were performed in electrospray positive ionization mode in the 50–750 m/z range in centroid mode. The parameters were as follows: spray voltage: 3 kV; sheath flow rate: 75; aux gas pressure: 20; capillary temperature: 350°C; heater temperature: 430°C. The analysis was conducted in Full MS data-dependent MS2 mode (Discovery mode). Resolution was set to 70,000 in Full MS mode, and the AGC (automatic gain control) target was set to 1x106. In MS2, resolution was 17,500, AGC target was set to 2x105, isolation window was 0.4 m/z, and normalized collision energy was stepped to 15, 30 and 40 eV. The UHPLC column was a Phenomenex Luna Omega Polar C18 1.6 μm, 150 x 2.1 mm. The column temperature was set to 42°C, and the flow rate was 0.4 mL min-1. The solvent system was a mixture of water (A) with increasing proportions of acetonitrile (B), with both solvents modified with 0.1% formic acid. The gradient was as follows: 1% B 3 min before injection, then from 1 to 15 min, a gradient increase of B up to 100% (curve 5), followed by 100% B for 5 min. The flow was injected into the mass spectrometer starting immediately after injection. All data were acquired and processed using FreeStyle 1.5 software (Thermo Fisher Scientific).

Chemicals and solvents. N-acyl-homoserine lactones (AHL) were obtained from Cayman Chemical (Ann Arbor, MI, USA). Stock solutions were obtained by dissolving standards in methanol or dichloromethane (C18-AHL) at a concentration of 1 mg mL-1 and stored at -80°C. Standard solutions for UHPLC-high-resolution tandem mass spectrometry (HRMS) analyses were prepared by diluting each individual standard solution with methanol in order to obtain a concentration range from 2000 to 20 ng mL-1. LC-MS grade methanol, acetonitrile and formic acid were purchased from Biosolve (Biosolve Chimie, Dieuze, France), analytical-grade ethyl acetate was obtained from Sigma-Aldrich. Pure water was obtained from Elga Purelab Flex System (Veolia LabWater STI, Antony, France).

Nature of lytic compounds secreted by V. atlanticus LGP32

To determine the nature of the lytic compounds secreted, V. atlanticus LGP32 grown for 60 h in Zobell media at 22 °C was filtered through a 10 kDa membrane (Amicon® Ultra-4 filter unit). The eluate containing molecules with MW below 10 KDa was then incubated in a water bath at 100°C for 30 min. Boiled filtrates (0.1 mL) were subsequently used to inoculate A. pacificum (ACT03 strain) cultures, then lytic activity was observed under the Leica TCS SPE confocal laser scanning system. Zobell media with the same treatment was used as control.

Statistical analysis

Environmental data. Statistical analyses were performed using R 3.6.3 software. The relationship between Alexandrium and Vibrio was explored separately in spring and autumn. We used a generalized linear model specifying a Gaussian family. For spring, the dataset for salinity and temperature was complete (14 periods of observation). One influential period was detected and removed from the dataset whenever no dead Alexandrium cells were observed. The effects of explanatory variables such as log10 (Vibrio+1), salinity and temperature were centred, reduced and tested as fixed effects with a linear relationship. Model selection was performed using the Akaike information criterion corrected for small sample size (AICc). Models were considered different whenever the difference between their AICc value and the lowest AICc value (ΔAICc) was lower than 2 (Burnham and Anderson, 2002). Alexandrium distribution and model residuals were checked for normal distribution assumptions (QQ plot and Shapiro-Wilk test). For autumn, the dataset was complete for 10 periods of observation. Salinity and temperature were missing for three periods. We explored the relationship between Alexandrium and Vibrio alone using the method detailed above for spring.

In vitro data. Statistical analyses were performed using one-way ANOVA (analysed by pair) followed by Tukey’s test (Statistica 10.0 software, StatSoft, Maison-Alfort, France). *P <0.05, **P <0.01, ***P <0.001

Results

Concomitant occurrence of A. pacificum ACT03, A. tamarense ATT07 and free-living Vibrio spp. in the Thau lagoon

In the spring and autumn of 2015, in the Thau Lagoon (Fig. 1A), we detected Alexandrium algae (A. pacificum ACT03 and A. tamarense ATT07, both alive and in degraded cell forms) and free-living Vibrio, but no plankton-associated Vibrio were observed (Fig. 1B, Data S1). Using model selection based on AICc, we found no significant relationship between Alexandrium (A. pacificum ACT03, A. tamarense) and Vibrio Spp. abundances in autumn. This result is consistent with the difficulty that Vibrio has in growing at temperatures below 20°C and with the complex interacting factors driving bloom dynamics (Laanaia et al., 2013). Interestingly, in spring 2015, the mean densities of Alexandrium and of free-living Vibrio were positively correlated. The lowest AICc was obtained with the model explaining degraded form of Alexandrium density based on the free Vibrio density (Fig. 1C). Given that, this model is not so different from the model with only the intercept, but better than any other linear combination with other potentially interfering drivers, such as temperature and salinity (Fig. 1D), we searched for evidence of a relationship between Vibrio and Alexandrium by studying their interaction in vitro.

Dynamics of Alexandrium and Vibrio in the environment.

(A) Location of the monitoring station in the Thau Lagoon (southern France). (B) Mean abundance (DNA equiv.) of Vibrio spp. (16S) and Alexandrium spp. (A. pacificum ACT03 + A. tamarense ATT07). Vibrio cells (black line with diamond dot) and degraded Alexandrium cells (grey line with round dot) were evidence in the 0.2–0.8 μm fraction (free Vibrio fraction) in spring and autumn 2015. Living Alexandrium cells (grey line with roundot) but no plankton-associated Vibrio spp. (black line with diamond dot) were evidence in the 0.8–180 μm in spring and autumn. (C) Result of Akaike information criterion (AICc) models tested to explain the mean value of degraded Alexandrium cells (dead cells) in spring. (D) Wald test of the AICc model attributing the mean value of degraded cells of Alexandrium in spring to free Vibrio.

Vibrio atlanticus LGP32 feeds on Alexandrium pacificum ACT03

To investigate whether Alexandrium interacts with Vibrio, we incubated in Enriched Natural SeaWater (ENSW) A. pacificum ACT03 with V. atlanticus LGP32 previously grown for 12 hours in Zobell media. In interaction with V. atlanticus LGP32, A. pacificum ACT03 cell abundance decreased significantly from 2.10 x 104 cells mL-1 to 1.07 x 104 cells mL-1 after 48 h (Fig. 2A), while the V. atlanticus LGP32 concentration grew significantly after 26 h of interaction, reaching a maximum peak density of 7.67 x 107 CFU mL-1 at 34 h (Fig. 2B). In the control experiment where A. pacificum ACT03 was cultured alone in ENSW, the algal concentration remained stable over time (Fig. 2A) and no bacteria were on the corresponding TCBS plates (Vibrio selective medium). In the control where V. atlanticus LGP32 was cultured alone in ENSW, the concentration of the bacteria tended to increase during the first 24 h (from 2.0 x 105 to 7.0 x 105 CFU mL-1) and then decreased at 48h, below the initial starting concentration (Fig. 2B). These results show that the interaction between V. atlanticus LGP32 and A. pacificum ACT03 leads to a decline in the algal population and promotes the growth of V. atlanticus LGP32. This suggests that V. atlanticus LGP32 is able to feed on A. pacificum ACT03.

Incubation of Vibrio atlanticus LGP32 and Alexandrium pacificum ACT03 in enriched natural seawater (ENSW).

(A) A. pacificum ACT03 cultured alone (grey bar) and incubated with V. atlanticus LGP32 (black bar) in ENSW. (B) V. atlanticus LGP32 cultured alone (grey bar) and incubated with A. pacificum ACT03 (black bar) in ENSW. (C) Images of the interaction of V. atlanticus LGP32-GFP with A. pacificum ACT03 taken at 8h00 and 28h00 of co-culture. V. atlanticus LGP32 (small green cells), living A. pacificum ACT03 cells (large red cells) and degraded A. pacificum ACT03 cells (large green cells). *P < 0.01 (analysis by pairs).

V. atlanticus LGP32 performs group-attacks on A. pacificum ACT03

Epifluorescence microscopy observation of GFP-labelled V. atlanticus LGP32 (previously grow in Zobell medium) in interaction showed that A. pacificum ACT03 cells that had lost their motility were attacked individually by groups of V. atlanticus LGP32 before being lysed (Movie 1). Attacks were extremely rapid, with empty thecae (algal envelopes) observed in the medium after less than 60 s (Movie 2). During the attack, V. atlanticus LGP32 did not invade the algal cell but remained clustered on the cell surface (Fig. 3A2).

Role of Vibrio atlanticus LGP32 starvation in the interspecific interaction process.

Experiments were conducted by incubating A. pacificum ACT03 with V. atlanticus LGP32 previously grown for 12, 36, 60 and 126 h in Zobell medium. (A) Percentage of motile A. pacificum ACT03. (B) A. pacificum ACT03 group-attacked by V. atlanticus LGP32 and (C) A. pacificum ACT03 lysis after 0, 15, 30, 45 and 60 min of interaction. Corresponding pictures showing (A1) Black arrows indicate unhooked and degrade flagellum from A. pacificum ACT03 flagellum, (B1) Image of the interaction of V. atlanticus LGP32-GFP with A. pacificum ACT03 taken at 8h00 of co-culture. V. atlanticus LGP32 (small green cells), living A. pacificum ACT03 (large red cells) and dead A. pacificum ACT03 (large green cell). (C1) Black arrow 1 indicate vesicle formation on A. pacificum ACT03 cell and black arrow 2 indicate exploded A. pacificum ACT03 cell. ***P <0.001.

Group attack of A. pacificum ACT03 is activated by V. atlanticus LGP32 starvation

In 2002, Martin hypothesized that nutritional stress induces bacteria to lyse algae (Martin, 2002). To test this hypothesis, we monitored V. atlanticus LGP32 behaviour in response to starvation (Fig. 3). We observed that V. atlanticus LGP32 in exponentially growth phase (12 h of culture in Zobell medium) did not interact with A. pacificum ACT03 cells for the first hour of contact (Fig. 3A, B, C). In contrast, V. atlanticus LGP32 in the decline phase (36 h of culture in Zobell medium) induced a significant decrease in the number of motile algae cells by 8.9% after 15 min and by 43.3% after 60 min (Fig. 3A). This phenomenon corresponded to the degradation and/or disruption of algal flagella (Movie 3). The flagella no longer functioned correctly, which caused erratic swimming of the algae (Video 3, left cell). This was followed by a complete cessation of swimming. When the flagellum detached from the algae (Video 3, right cell), the attack occurred. With starved V. atlanticus LGP32 (60 h of culture in Zobell medium), algae immobilization was fast and significant (91.4% in 15 min, Fig. 3A), and algae were attacked individually being targeted by groups of V. atlanticus LGP32 (Movie 4). The percentage of cells attacked and killed peaked at 30% after 15–30 min of contact (Fig. 3B) and then decreased. After 1 h, attacks had stopped with approximately 40% of the algal cells still alive (Fig. 3C). Although it remains unclear whether the attacks occur during a specific phase of growth, it is evident that the cells are already weakened before attack as they have all lost their flagella. An old-starved culture of V. atlanticus LGP32 (126 h) significantly immobilized A. pacificum ACT03 cells within a few minutes, with lysis occurring immediately (Fig. 3A, C), making it impossible to detect attacks by V. atlanticus LGP32 (Fig. 3B). The lysis phase corresponded to initial vesicle formation followed by the bursting of A. pacificum ACT03 cells (Movie 5) and was induced by the old-starved culture supernatant of V. atlanticus LGP32 (Fig. S1).

We next tested whether this lytic effect was mediated by thermostable molecule (s) secreted by Vibrio. The culture supernatant of starved culture of V. atlanticus LGP32 (36 h) filtered through a 10 kDa membrane and then incubated at 100°C for 30 min still possessed its lytic properties, indicating that the algicidal compounds produced by V. atlanticus LGP32 are small thermostable molecules unlikely to be lytic enzymes, or lysins able to digest the algae cell. However, these experimental observations clearly show the key role of nutrient limitation in triggering the ‘group-attack’ behaviour and the secretion of lytic compounds of V. atlanticus LGP32.

Group attack occurs on A. pacificum ACT03 in exponential phase of growth

Here, we wondered whether the live/dead status of algae is important for V. atlanticus LGP32-mediated group attacks targeting. To this aim, A. pacificum ACT03 in exponential growth phase was first exposed for 30 min to supernatant from 60 hours starved V. atlanticus LGP32 Zobell media that induced 25% lysis of A. pacificum ACT03 cells and next to the corresponding V. atlanticus LGP32 cells. Group attacks were observed on non-degraded A. pacificum ACT03 cells, but not on lysed cells. This result is similar to what is observed on the Movie 1 with flash attacks only on immobilized, but not degraded A. pacificum ACT03 cells (red), and not on lysed cells (green). In addition, no group attacks occurred on cells from an old A. pacificum ACT03 culture (1-month culture). Together, with the very short duration of group attacks (Movie 1), these results indicate that V. atlanticus LGP32 group-attacked exponentially growing cells of A. pacificum ACT03, but not decomposing algae, suggesting that this behaviour is not just an opportunistic response of heterotrophic bacteria to organic substrates.

Group attack is independent of quorum sensing

Considering the coordinated action of V. atlanticus LGP32 in attacks on A. pacificum ACT03, we tested whether the group-attack process depended on the key physiological mechanism that regulates many functions in marine microbial cells, quorum sensing (QS) (Lami, 2019; Papenfort and Bassler, 2016), a type of cell-cell communication. Although QS is a cell-density-dependent mechanism, our results showed no attack from a 12 h culture of V. atlanticus LGP32 up to a concentration of 4.106 Vibrio mL-1 (Fig 4A). Attacks were only observed with V. atlanticus LGP32 from a 60 h culture at low concentration of 5.103 Vibrio mL-1 to the highest concentration tested of 5.105 Vibrio mL-1 (Fig 4A), consistent with the hypothesis that the attacks were independent of Vibrio density.

Role of quorum sensing and the vibrioferrin iron uptake pathway in the interaction process.

(A) Effect of V. atlanticus LGP32 cell density on the predation process. Experiments examined the impact of V. atlanticus LGP32 cell density on its ability to attack A. pacificum ACT03. A. pacificum ACT03 cells (2.104 cells) were incubated with V. atlanticus LGP32 grown for 60 hours in Zobell medium at concentrations ranging from 5×103 to 5×105 cells mL−1 (represented by black bars). For comparison, A. pacificum ACT03 was also incubated with V. atlanticus LGP32 grown for only 12 hours in Zobell medium at concentrations ranging from 5×103 to 4×106 cells mL−1 (represented by grey bars). The term “Att” indicates cells of A. pacificum ACT03 that were subjected to group attacks by V. atlanticus LGP32. (B) CqsS, luxM, luxN, luxS, and luxP quorum sensing and PvsA, PvuB and PvuA2 vibrioferrin pathway genes expression in V. atlanticus LGP32 grown for 12, 36 and 60 h in Zobell medium. (C) Effect of V. atlanticus LGP32 mutants on the group-attacked process. Experiments were conducted by incubating A. pacificum ACT03 with V. atlanticus LGP32, V. atlanticus LGP32 tagged with GFP, V. atlanticus LGP32 washed with ENSW or V. atlanticus LGP32 mutant ΔPvuB, ΔluxM, ΔluxR and ΔluxS previously grown 60 h in Zobell media (control), The percentage of A. pacificum ACT03 group-attacked was determined during the first 30 min of exposure. (D) Effect of V. atlanticus LGP32 cultures media composition on the group-attacked process. Experiments were conducted by incubating A. pacificum ACT03 with V. atlanticus LGP32 grown 60 h in Zobell media supplemented with H3BO4 or FeCl3. The results were compared with an exposure to V. atlanticus LGP32 grown 60 h in Zobell media. **P <0.01, ***P <0.001

The analysis of the expression of genes involved in the known QS pathways in Vibrio cell (Fig. S3A), highlighted that only the AI-2 pathway was induced during nutrient stress of V. atlanticus LGP32, because only the expression of the AI synthase (LuxS) and its receptor (LuxP) increased significantly (Fig. 4B; ANOVA p <0.05). This was confirmed by a QS bioluminescence assay, which showed a AI-2 molecules (unquantified) in the Zobell culture supernatant of V. atlanticus LGP32 after 60 h of culture but not after 12 h of culture and not in the ENSW supernatant of V. atlanticus LGP32 after 12 or 60 h of culture. UHPLC-HRMS/MS provided no evidence of detectable HAI-1) and CAI-1 in any experiments. By targeted mutagenesis of key genes involved in QS pathways ΔluxM (HAI-1 production), ΔluxS (AI-2 production) and ΔluxR (high-density QS master regulator) did not lead to any change in the group-attack behaviour of V. atlanticus LGP32 (Fig. 4C). Taken together these results showed that group-attack by V. atlanticus LGP32 is not coordinated by QS.

Group attack related to the availability of iron

The comparative analysis of the proteome of V. atlanticus LGP32 incubated 60 h in artificial seawater (ENSW) versus V. atlanticus LGP32 grown 12 h in Zobell nutrient-rich medium revealed 10 proteins modulated by nutrient stress (Fig. S2). The two most down-regulated proteins correspond to be ß-ketoacyl-(acyl-carrier-protein) synthase II (-22-fold in ENSW compared to Zobell), a key regulator of bacterial fatty acid synthesis, and the dihydroorotase (-6.6-fold in ENSW compared to Zobell), an enzyme essential for pyrimidine biosynthesis and thus bacterial proliferation and growth. The low expression of these proteins in ENSW is consistent with V. atlanticus LGP32 nutritional starvation. The most up-regulated protein in starved V. atlanticus LGP32, with an increase of more than 6-fold, was glucosamine-6-phophate deaminase, an enzyme involved in bacterial energy metabolism probably necessary for its survival. Among the other up -regulated proteins, one was an iron siderophore-binding protein (Spot 4413, Fig. S2A) corresponding to the vibrioferrin outer membrane receptor PvuB, whose gene is part of the pvu operons involved in iron transport (Fig. S3B). Interestingly, the corresponding gene pvuB as well as the vibrioferrin membrane receptor gene pvuA2 (Fig. S3B) were both significantly induced in Vibrio under nutrient stress (Fig. 4B; ANOVA p <0.01) but not the one involved in the vibrioferrin biosynthesis, pvsA (Fig. 4B). Remarkably, among the 10 proteins identified by proteomic analysis only V. atlanticus LGP32 mutant lacking pvuB failed to attack A. pacificum ACT03 (Fig. 4C; ANOVA p <0.001). In the absence of the pvuB gene, V. atlanticus LGP32 was unable to attack in group A. pacificum ACT03. In addition, V. atlanticus LGP32 cells that had been washed with ENSW to remove their culture supernatant metabolites also failed to attack A. pacificum ACT03 (Fig. 4C; ANOVA p <0.001), which is congruent with the hypothesis that attacks depend on the V. atlanticus LGP32 vibrioferrin transport system. Finally, group-attacks increased significantly, when FeCl3 was added to the Vibrio culture medium (Fig. 4D) but not with H3BO4 (Fig. 4D). Taken together, those results are consistent with the hypothesis that attacks are regulated by iron.

Group attack is a Vibrio spp. behavior specific to Alexandrium spp

To evaluate the dinoflagellates specificity of the group-attack behaviour, a selection of Vibrio spp. was co-cultured with a selection of dinoflagellate strains commonly found in the Mediterranean Sea. The results showed that, among the Vibrio spp. tested (pathogenic or not) all, under nutrient stress, were able to secrete algicidal compounds, immobilize, ‘group-attack’ and lyse A. pacificum ACT03 cells (Table 1) and no tropism linked to their pathogenesis for fish of invertebrates was observed (Table 1). Among the sixteen dinoflagellates species tested, all eight Alexandrium spp. (non-toxic and paralytic shellfish toxin (PST) producers) and Gymnodinium catenatum (PST producer) were immobilized, ‘group-attacked’ and lysed by V. atlanticus LGP32 (Table 2), but no tropism linked to PSTs was revealed (Table 2).

Discussion

Predation is a widespread mode of interaction for survival in the natural world (Finke and Denno, 2004; Sinclair et al., 2003). Among predators, predatory bacteria are found in a wide variety of environments, and like bacteriophages and predatory protists, they have been reported to prey exclusively on other bacteria (Johnke et al., 2014; Perez et al., 2016). Considering predator as a free organism that feeds at the expense of another, this study is the first evidence of the capacity of some Vibrio to develop a predatory strategy against an alga. This behaviour differs from parasitism, because the survival of Vibrio is not exclusively dependent on the algae.

The strategy developed by Vibrio to kill algae may be reminiscent of strategies previously described in the prokaryotes (Johnke et al., 2014). As shown in Movie 1, the interaction between V. atlanticus LGP32 and A. pacificum ACT03 proceeds in three stages (Fig. 5). The first stage, the ‘immobilization stage’, recalls the strategy used by Streptomyces to immobilize its prey (Kumbhar et al., 2014) based on the secretion of algicidal metabolites that disrupt the flagella. The second stage, the ‘attack stage’, resembles the ‘wolf-pack attack’ strategy described for Myxococcus xanthus and Lysobacter (Martin, 2002; Perez et al., 2014). V. atlanticus LGP32 also surrounds A. pacificum ACT03 cells at high density for a very short time, but does not invade the algal cell. This phenomenon is comparable to that of bacteria clustering around lysed ciliate cells (Blackburn et al., 1998). The third stage, the ‘killing stage’, is similar to that of epibiotic bacterial predators, which induces the lysis of bacteria from the outside (Rashidan and Bird, 2001). Overall, these observations show that V. atlanticus LGP32 is able of wolf-pack hunting behaviour. Results showed that quorum sensing does not control algal predation by Vibrio, in contrast to iron acquisition systems, which play a major role in regulating this phenotype, as we demonstrated genetically. Indeed, our quorum sensing mutants exhibited the same predatory phenotype as the wild-type Vibrio atlanticus LGP32, whereas mutants impaired in iron uptake completely lost this phenotype. However, quorum sensing and iron acquisition are sometimes interconnected in Vibrio (McRose et al., 2018). For instance, in Vibrio vulnificus, the production of vulnibactin (a siderophore) is controlled by AI-2 (Kim and Shin, 2011). Whether the production of AI-2 by V. atlanticus LGP32, induced during group attacks of A. pacificum ACT03, promotes the production of vibrioferrin (Lami, 2019) remains to be investigated.

Schematic representation of a putative strategy developed by Vibrio spp. to feed on Alexandrium spp. and G. catenatum in the environment.

(A) Vibrio in the environment when subjected to starvation secrete non-protein lytic compounds. (B) Some of these lytic compounds degrade the flagella, immobilizing the alga (immobilization stage). (C) Then Vibrio swims and clusters around its prey coordinating group attacks (group-attack stage). (D) Lytic compounds released by Vibrio where able to concentrate around the algae cells, thereby lysing the algae (killing stage). (E) Feeding on the released nutrients, Vibrio multiply and then spread in the environment. Yellow clouds: Lytic compound release by Vibrio, Grey clouds: Algal nutrients released upon lysis.

We further show that the group-attack behaviour of V. atlanticus LGP32 depends on iron availability. Iron is an essential element for growth in most organisms, including phytoplankton (Martin and Fitzwater, 1988) and bacteria (Neilands, 1981), and its concentration in seawater is known to be very low with measurements in the open ocean surface at 0.2 nM and in deep waters at 0.6 nM (Millero, 1998).

Moreover, Its low solubility in seawater limits its availability (Bruland et al., 1994; Wu and Luther, 1994). To acquire iron, bacteria have developed systems based on the secretion (and subsequent uptake) of iron-chelating siderophores to obtain this element from the environment (Amin et al., 2009a). Therefore, many Vibrio spp. produce a siderophore known as vibrioferrin, which is synthesized and secreted by proteins encoded by the pvsABCDE gene cluster (Fig. S3). Boron being known to be a regulator or capable of being transported by vibrioferrin (Romano et al., 2013; Weerasinghe et al., 2013) we tested its potential involvement in the interaction but no effect was evidenced here. For iron-vibrioferrin uptake, Vibrio parahaemolyticus uses a membrane siderophore receptor, called PvuA, which is coupled to an inner membrane ATP-binding cassette (ABC). This ABC transporter system comprised of proteins encoded by the pvuABCDE gene cluster (Fig. S3) is required for transporting the siderophore across the inner membrane (Tanabe et al., 2003). Siderophores are not only iron carriers but also important regulators of virulence (Miethke and Marahiel, 2007) and mediators of bacterial interaction with phytoplankton (Amin et al., 2009b; Kramer et al., 2020). We showed here a pivotal role of iron in the interaction between V. altlanticus LGP32 and A. pacificum ACT03. This mirrors the mutualistic interaction observed between Gymnodinium catenatum and Marinobacter (Amin et al., 2009b). In fact, in natural settings, the co-occurence of Marinobacter and G. catenatum is suggested to depend on a mutually beneficial utilization of iron and carbon resources (Bolch et al., 2011). As in the present study, iron seems to play a key role in the interaction. Indeed, the labile iron released through the photolysis of ferric chelates with vibrioferrin providing a crucial iron source for phytoplankton, which need substantial amounts of iron to support carbon fixation through photosynthesis (Amin et al., 2009a; Yang et al., 2021). This fixed carbon, in turn, sustains the growth of both the phytoplankton and their bacterial counterparts (Amin et al., 2009b; Kramer et al., 2020). Future study could demonstrate further the role of vibrioferrin in group attack, by adding iron-saturated vibrioferrin to algae-Vibrio co-cultures.

In the natural environment, associations between bacteria and algae have already been observed (Lopez-Joven et al., 2018; Miller et al., 2005; Rosales et al., 2022; Xu et al., 2022). Although we do not have definitive proof of a predator/prey interaction in situ due to confounding environmental variables, environmental data acquired in the Thau Lagoon showed a correlation between Alexandrium and Vibrio occurrence, suggesting that such interaction could occur in nature. Evolution of such a predatory behavior would provide an important ecological advantage to Vibrio to obtain nutrients in environment, where Alexandrium spp. and Gymnodinium catenatum form blooms.

With more than 30 species distributed all over the world (Anderson et al., 2012; Hallegraeff et al., 2021), Alexandrium spp. and Gymnodinium spp. considered as invasive species by the Delivering Alien Invasive Species Inventories for Europe (http://www.europe-aliens.org)), could play an unexpected and important role in maintaining, structuring and regulating Vibrio populations in the ecosystem. In turn, Vibrio could contribute to the regulation and control of their blooms.

To conclude, this study reveals the capacity of some Vibrio spp. to be facultative predators that hunt specific algae. In the current context of climate change favourable to their development monitoring the invasive algae, Alexandrium spp. and Gymnodinium catenatum should be considered not only for their potent harmful effect on humans and animals, but also as a potential source of nutrients for the expansion of Vibrio, particularly pathogenic species (Lemire et al., 2015).

eLife Assessment

This important study convincingly shows that Vibrio bacteria act as predators of ecologically significant algae that contribute to harmful blooms in the lab, as well as in their natural habitat. While the data strongly suggest that starvation may induce predation, further work is needed to fully establish this link. Similarly, the evidence for a social component in the predation process remains incomplete. This study will be very impactful to those interested in the diversity of microbial predator-prey interactions and controlling toxic algal bloom, but the paper could be strengthened by more clearly showing the degree of replication, by better defining the terms used to describe the observed behaviour, and by providing better support for starvation and collective behaviour.

Reviewer #1 (Public review):

Summary:

Rolland and colleagues investigated the interaction between Vibrio bacteria and Alexandrium algae. The authors found a correlation between the abundance of the two in the Thau Lagoon and observed in the laboratory that Vibrio grows to higher numbers in the presence of the algae than in monoculture. Time-lapse imaging of Alexandrium in coculture with Vibrio enabled the authors to observe Vibrio bacteria in proximity to the algae and subsequent algae death. The authors further determine the mechanism of the interaction between the two and point out similarities between the observed phenotypes and predator-prey behaviours across organisms.

Strengths:

The study combines field work with mechanistic studies in the laboratory and uses a wide array of techniques ranging from co-cultivation experiments to genetic engineering, microscopy and proteomics. Further, the authors test multiple Vibrio and Alexandria species and claim a wide spread of the observed phenotypes.

Weaknesses:

In my view, the presentation of the data is in some cases not ideal. The phrasing of some conclusions (e.g., group-attacks and wolf-pack-hunting by the bacteria) is in my opinion too strong based on the herein provided data.

Reviewer #2 (Public review):

Goal summary:

The authors sought to (i) demonstrate correlations between the dynamics of the dinoflagellate Alexandrium pacificum and the bacterim Vibrio atlanticus in natural populations, ii) demonstrate the occurrence of predation in laboratory experiments, iii) claim coordinated action by the predators in the predation process, iv) demonstrate that predation is induced by predator starvation, and v) test for effects of quorum sensing and iron-uptake genes on the predation process.

Strengths include:

(1) Data indicating correlated dynamics in a natural environment that increase the motivation for the study of in vitro interactions.

(2) Experimental design allowing clear inference of predation based on population counts of both prey and predators in addition to microscopy-based evidence.

(3) Supplementation of population-level data with molecular approaches to test hypotheses regarding possible involvement of quorum sensing and iron uptake in predation.

Weaknesses include:

(1) A lack of early, clear definitions for several important terms used in the paper, including 'predation', 'coordination' and 'coordinated action', 'group attack', and 'wolf-pack hunting', along with a corresponding lack of criteria for what evidence would warrant use of some of these labels. (For example, does mere simultaneity of attacks of an A. pacificum cell by many V. atlanticus cells constitute "coordination"? Or, as it seems to us, does coordination require some form of signalling between predator cells?)

(2) Absence of controls for cell density in the test for starvation effects on predatory behavior; unclear how the length of incubation affects the density of V. atlanticus cells.

(3) Lack of clarity in some of the methodological descriptions

Appraisal:

The authors convincingly achieve their aim of demonstrating that V. atlanticus can prey on A. pacificum, provide strongly suggestive evidence that such predation is induced by starvation, and clearly demonstrate that both iron availability and, correspondingly, the presence of genes involved in iron uptake, strongly influence the efficacy of predation. However, the evidence for starvation-induction of predation can be strengthened with cell-density controls; evidence for a social component to predation - positive interactions between attacking predators - is lacking.

Discussion of impact:

This paper will interest those interested in how microbial behaviour responds to environmental fluctuations, in particular predatory behaviour, but will do so more strongly if the evidence of starvation-induction of predation is strengthened. It will also interest those investigating bacteria-algae interactions and potential ecological controls of algal blooms. It has the potential to interest researchers of microbial cooperation, should the authors be able to provide any evidence of coordination between predator cells.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public review):

Weaknesses:

In my view, the presentation of the data is in some cases not ideal. The phrasing of some conclusions (e.g., group-attacks and wolf-pack-hunting by the bacteria) is in my opinion too strong based on the herein provided data.

We agree with your comment and have replaced the terms “Group-attacks” and “wolf-pack-hunting by “attacks” throughout the manuscript.

Reviewer #1 (Recommendations for the authors):

(1) Figure 2AB, please add the name of the statistical test and the number of replicates that the data is based on to the figure legend.

We thank Reviewer#1 for highlighting the need for more detail. We have revised the manuscript accordingly. The captions of figures 2, 3, 4 and S1 were revised to include the name of the statistical test and the number of replicates. Asterisks indicate significant differences in a multiple comparison test (One -way ANOVA with post hoc Tukey test),* P ≤ 0.05, ** P≤0.01, *** P≤ 0.001

(2) Figure 2C is this figure referred to in the text?

We apologize for this oversight. Figure 2C was replaced by new figures 2C and 2D and the old figure 2C is now referenced in the manuscript as Fig 3B1.

(3) Movie 1, could the movie please also be provided as .mp4? I suggest including individual images across time in the main figure so that readers do not rely on opening a supplementary file for this key finding of the study.

In the revised manuscript, all the videos were converted to mp4 format and individual images across time were included in Figure 2C and 2D (Chronological snapshots of one attack) and in figure 3B1 (Chronological snapshots of the complete event), thereby improving the readability of the manuscript.

(4) Figure 3A2 (text l. 355), I am afraid I do not find this figure.

Fig. 3A2 which previously corresponded to Fig. 3B1, correspond now to Fig. 2C and Fig. 2D. This has been corrected in the revised version of the manuscript.

(5) Lines 356ff, I am afraid that I find it hard to follow what the authors refer to as the right cell or the left cell. I suggest either adding labels to the movies or providing individual images across multiple timepoints into the main figure that can be labelled and bring across the point.

Arrows have been added to videos 3–5 to clearly indicate the cells referred to in the text and facilitate tracking across time.

(6) In general, for all the microscopy, on how many cells have these phenomena been observed? What is n=x? Has this been quantified?

We thank the reviewer for pointing this out.

In caption of Fig. 3, the sentence “(A) Percentage of motile A. pacificum ACT03. (B) A. pacificum ACT03 attacked by V. atlanticus LGP32 and (C) A. pacificum ACT03 lysis after 0, 15, 30, 45 and 60 min of interaction. “was replaced by “(A) Cumulative percentage of motile A. pacificum ACT03 cells. (B) Cumulative number of cells attacked by V. atlanticus LGP32 and (C) Cumulative cell lysis after 0, 15, 30, 45 and 60 minutes of interaction.”. In Fig. 3 caption, the sentence “All percentages were determined based on a minimum of 2,000 cells of A. pacificum ACT03.” was also added.

In Fig. 4 caption, the sentence “All percentages were determined based on a minimum of 2,000 cells of A. pacificum ACT03.” was added.

In Fig. S1 caption, the sentence “All percentages were determined based on a minimum of 2,000 cells of A. pacificum ACT03.” was added.

(7) Figure S1A, does this figure show means plus/minus standard deviation? If yes, please add this to the figure legends.

In Fig. S1 caption, the sentence “Error bars represent the standard deviation of the mean of three independent experiments” was added.

How do the authors explain the big variation in the test condition and not in the control?

Regarding the higher variation observed in the test condition compared to the control, this may, on the one hand, reflect biological variability between independent batches of 60-h V. atlanticus cultures used to prepare the supernatants, and, on the other hand, a heterogeneity in the physiological status of independent algal batches (N = 3 ; 2 × 10^4 cells ; see Materials and Methods, Co-culture assay), which may not be perfectly synchronized . In contrast, the control condition consists of A. pacificum cultures incubated in fresh medium without bacterial supernatant, for which algal motility is highly reproducible and thus shows very little variation.

(8) Line 375, "The lysis phase corresponded to initial vesicle formation followed by the bursting of A. pacificum ACT03 cells (Movie 5) and was induced by the old-starved culture supernatant of V. atlanticus LGP32 (Fig. S1)." Is this reference to Figure S1 correct? S1 shows motility, doesn't it? I don't see how this data supports the statement made in this sentence.

We apologize for this unclear message.

"The lysis phase corresponded to initial vesicle formation followed by the bursting of A. pacificum ACT03 cells (Video 5) and was induced by the old-starved culture supernatant of V. atlanticus LGP32 (Fig. S1)." was replaced by "The lysis phase corresponded to initial vesicle formation followed by the bursting of A. pacificum ACT03 cells (Fig. 3C and 3C1).

And “We next tested whether this lytic effect was mediated by thermostable molecule (s) secreted by Vibrio. “was replaced by “We next tested whether this lytic effect was linked to Vibrio culture supernatant and mediated by thermostable molecule (s) secreted by Vibrio.

(9) Line 388ff, "Group attacks were observed on non-degraded A. pacificum ACT03 cells, but not on previously lysed cells." No reference to a figure is provided. I am afraid I don't see the data that this statement is based on.

As it is impossible to show a lack of attack, we just clarified the basis of our experiment.

“To this end, A. pacificum ACT03 in exponential growth phase was first exposed for 30 minutes to the supernatant of a 60-hour culture of V. atlanticus LGP32, which induced 25% lysis of A. pacificum ACT03 cells. Next, the corresponding V. atlanticus LGP32 cells were added. During exposure, attacks were observed only on undegraded A. pacificum ACT03 cells, but not on previously lysed cells” was replaced by “To this end, A. pacificum ACT03 in exponential growth phase was first exposed for 30 minutes to the supernatant of a 126-hour culture of V. atlanticus LGP32, which induced lysis of 70% of the A. pacificum ACT03 cells (Figures 3C and 3C1, arrow 2 and video 4). Next, cells of V. atlanticus LGP32 from a 60-hour culture, capable of attacking A. pacificum ACT03 cells (Fig. 3B), were added. For 1 hour of exposure, no attack was observed on the previously lysed algae.”

(10) Figure 4a, Based on the labeling of the figure, in particular the x-axis, it is not fully clear to me what I am looking at.

Figure 4A has been reworked and its legend modified. We hope that this graph is clearer now.

(11) Line 428, did the authors consider complementing the pvuD deletion mutant and testing for gain of function when providing the gene in trans?

We did not investigate pvuD in this study and did not construct a pvuD deletion mutant. We therefore assume that the recommendation refers to pvuB, which was the focus of our work. Unfortunately, we did not perform this experiment. However, several lines of evidence support the implication of PvuB and the vibrioferrin uptake system in this process: (i) the loss of attack behaviour is specific to the mutant in the vibrioferrin uptake pathway and (ii) our expression and proteomic data show a strong induction of vibrioferrin uptake components under starvation and iron-manipulated conditions, which correlate with the attack phenotype.

(12) Use of the term "group attack" in parentheses in the text, but in the section header and title. Is there really sufficient actual data to say that this is a "group attack"? What exactly are the indications for this being a behaviour of a group?

We agree with you. The terms “group attacks” and “wolf-pack hunting” were replaced by the more neutral term “attacks” throughout the manuscript.

(13) Table S1 and S2, those tables give a nice overview. Do the authors provide the raw data based on which they make a claim on "+" and "-" in the individual categories? I would prefer to see the actual data or at least have the possibility to look into this.

In the revised versions of Tables 1 and 2, we have improved the captions and clarified the meaning of each column in order to avoid any ambiguity between the results of this study and the bibliographic information.

Specifically regarding Table 2 :

We do not present any visuals of the interaction between Vibrio and Alexandrium because these species all look alike. Regarding the other algae species tested in interaction with Vibrio, phenomena other than lysis or cell attack have been observed and are the subject of specific laboratory studies.

(14) Line 456 "first study", line 40f "first evidence of a new mechanism". I suggest toning this down a bit and being clearer in the abstract about this being a working model that can be suggested based on individual bits of data.

We thank Reviewer #1 for this helpful suggestion.

In the summary:

“This is the first evidence of a new mechanism that could to be involved in regulating Alexandrium spp. blooms and giving Vibrio a competitive advantage in obtaining nutrients from the environment.” was replaced by “The interaction model we propose here suggests that Vibrio could play a role in regulating the proliferation of Alexandrium spp., giving it a competitive advantage in obtaining nutrients from the environment.”

In the discussion:

Considering predator as a free organism that feeds at the expense of another, this study is the first evidence of the capacity of some Vibrio to develop a predatory strategy against an alga. This behaviour differs from parasitism, because the survival of Vibrio is not exclusively dependent on algae in environment” was replaced by “Consider a predator as a free-living organism that kills its prey and feeds on it, this study provides data suggesting the ability of Vibrios to develop an original predator-like behaviour to kill and feed on algae.”

(15) Line 469 "Overall, these observations show that V. atlanticus LGP32 is able of wolf-pack hunting behaviour." I see the similarities. I feel that the term "show" is a bit too strong here, or I suggest referring to "wolf-pack-like behaviour".

The sentence “Overall, these observations show that V. atlanticus LGP32 is able of wolf-pack hunting attack behaviour” was replaced by “Overall, these observations suggest that V. atlanticus LGP32 can exhibit a predator-like behaviour”

Reviewer #2 (Public review):

As Weaknesses Reviewer #2 include:

(1) A lack of early, clear definitions for several important terms used in the paper, including 'predation', 'coordination' and 'coordinated action', 'group attack', and 'wolf-pack hunting', along with a corresponding lack of criteria for what evidence would warrant use of some of these labels. (For example, does mere simultaneity of attacks of an A. pacificum cell by many V. atlanticus cells constitute "coordination"? Or, as it seems to us, does coordination require some form of signalling between predator cells?)

The term “Coordinate” was replaced by “simultaneous” throughout the manuscript

The terms “Group attack” and “wolf pack hunting” were replaced by “attack” throughout the manuscript

(2) Absence of controls for cell density in the test for starvation effects on predatory behaviour; unclear how the length of incubation affects the density of V. atlanticus cells.

We thank the reviewer for pointing this out.

Cells density experiment was already performed (cf. Fig. 4A).

The sentence. ”All percentages were determined based on a minimum of 2,000 cells of A. pacificum ACT03.“ was added in captions of Fig. 3, Fig. 4 and Fig S1

(3) Lack of clarity in some of the methodological descriptions

The Methodology has been checked and some improvements have been made.

Reviewer #2 (Recommendations for the authors):

(A) Title

(1) Could 'induces' be better than 'promotes'?

We agree with Reviewer #2. The initial title, “Starvation of the bacterium Vibrio atlanticus promotes lightning group-attacks on the dinoflagellate Alexandrium pacificum”, was replaced by “Starvation of the bacterium Vibrio atlanticus induces simultaneous attacks on the dinoflagellate Alexandrium pacificum”.

(B) Abstract

(1) Perhaps define pycosphere in the abstract - many readers might not know this word.

We have revised the abstract to define the term phycosphere and added the sentence “This occurs in the microenvironment surrounding phytoplankton cells, the phycosphere. An interface rich in nutrients and organic molecules exuded by the cell.”

(2) Perhaps "on dinoflagellates".

We thank Reviewer #2 for this suggestion. We have revised the abstract by replacing “on the dinoflagellates species” with “on dinoflagellates”.

(3) Line 33 - The word 'prey' is used without a claim of predation having yet been made; only killing has been claimed so far.

We agree and have replaced the word “prey” by “algae” in the abstract.

(4) Line 34 - It is unclear whether the description refers to the 'attack stage' or to 'wolf-pack attack' in general. The sentence is written in such a way that it seems to refer to 'wolf-pack attack'. However, this would seem to be incorrect, with the description being specific to V. atlanticus.

To avoid this ambiguity, we have removed the sentence “resembles the ‘wolf-pack attack’ strategy” from the abstract.

(5) Line 35 - Should there be a 'consumption phase'?

We agree with the reviewer #2, “degradation” was replaced by “consumption”.

(6) If predation is claimed later in the manuscript (which it is), it should be explicitly claimed in the abstract.

We thank Reviewer #2 for this helpful suggestion.

We have revised the abstract. The sentence “Results showed that Vibrio atlanticus was able to coordinate lightning group attacks then kill the dinoflagellate Alexandrium pacificum ACT03” was replaced by “The results showed that Vibrio atlanticus was capable of attacking and killing the dinoflagellate Alexandrium pacificum ACT03”.

(C) Main text

(1) Line 54 - Perhaps "Among HAB-causing organisms...".

We agree with the reviewer’s suggestion and have revised the wording.

(2) Line 56 - "that, together with..., form the "Alexandrium tamarense" complex".

We agree with the reviewer’s suggestion and have revised the sentence.

(3) Line 57 - What this "complex" is and its significance should be explained.

“Among them, Alexandrium pacificum is a flagellated eukaryotic unicellular organism that together with Alexandrium tamarense and Alexandrium fundyense form the "Alexandrium tamarense" complex (Hadjadji et al., 2020)” was replaced by

“Among them, Alexandrium pacificum is a flagellated eukaryotic unicellular organism that together with Alexandrium tamarense and Alexandrium fundyense form the "Alexandrium tamarense" complex, responsible for paralytic shellfish poisoning worldwide (Hadjadji et al., 2020)”

(4) Line 58 - What is a Rephy survey?

We clarified this point, “by rephy survey” was replaced by “by the French phytoplankton observation and monitoring network (Rephy)”

(5) Line 59 - 'resulting in' instead of 'resulting of'.

We agree with the reviewer and have replaced “resulting of” with “resulting in”.

(6) Line 65 - It seems that ', influencing the time of appearance of blooms' would be more correct than the current phrasing. The current phrasing is unclear regarding the relation between species, tolerance range, and the time of appearance of blooms.

To address this point, “Depending on the phytoplankton species, the tolerance range of physicochemical parameters is different and influences the time of appearance of blooms” was replaced by “Depending on the species of phytoplankton, tolerance to physicochemical parameters varies, which influences when blooms occur.”

(7) Line 76 - Run-on sentence which should probably be split after the reference to Wang et al., 2020.

We agree with the reviewer and have split the sentence.

(8) Line 89 - What are these observations?

This sentence was reformulated.

“Based on observations from the natural environment showing a potent relationship between Vibrio and Alexandrium algae bloom events, this study aim to determine in vitro, the main factors implicated in this relationship” was replaced by ”This study aims to describe observations made in the natural environment between Vibrio bacteria and Alexandrium algal blooms, and to determine in vitro the main factors involved in this relationship.”

(9) Line 94 - This is the first clear reference to a predator-prey interaction, and it is stated as if it's established. Is it not a central goal of the study to demonstrate that predation is even happening?

Based on the title and abstract, I would have expected the major claims of the paper highlighted in the abstract to be:

(i) that predation of algae by bacteria occurs in this system,

(ii) there is a social component of predation,

(iii) claims about what induces this predatory behaviour.

The summary has been amended accordingly, and the term “predation” has been removed, along with all sentences referring to it.

(10) Line 99 - What does n.d. mean?

This point was addressed in the revised version.

(11) Line 97 section - specify qPCR.

This point was clarified in the revised version.

(12) Line 139 - Mentioning the oligonucleotides in this part of the methods seems out of place. Would this not fit better in the section on Gene expression analysis?

This sentence was discarded from this paragraph.

(13) Line 147 - Where did the co-cultured phytoplankton species come from?

To answer this point, reference to Table 2 was added

(14) Line 149 - Is it known if the phytoplankton strains had all grown to the same density after 24 hours?

The doubling time of dinoflagellates in laboratory culture is between 5 and 7 days. During the duration of the experiments, the dinoflagellate concentration did not change significantly.

The sentence “(doubling time between 5 and 7 days)” was added

(15) Line 150 - Was the density of the Vibrio cultures at the different incubation times measured? Density might play an important role in predation, and so it would be important to control for density in these assays.

The concentrations of live vibrio in each individual culture were not actually measured. However, the role of vibrio density in attacks was measured and is shown in Figure 4A and observed in Fig 2B.

(16) Line 153 - How long was the co-incubation?

The incubation times were added in the revised version.

(17) Line 158 - What is mean by "independent experiments", more exactly?

To clarify this point, “Data are the means of three independent experiments” was replaced by “The data come from three independent experiments using independent phytoplankton cultures and independent bacterial cultures.”

(18) Line 161 - Perhaps give the source information about the Vibrio strain at its first mention.

A reference has been added in the revised preprint.

(19) Line 163 - line 141 refer to multiple non-axenic species, whereas here "the algal strain" is referred to.

And

(20) Line 164 - language phrasing throughout the manuscript could use some polishing, e.g., "this means that additional bacteria...".

To address this comment, “As the algal strain used in the study is not axenic, means that additional bacteria, other than the V. atlanticus LGP32, are potentially present in the experiments.” was replaced by “As the A. pacificum ACT03 strain (table 2) used in the study is not axenic, there is potential for bacteria other than V. atlanticus LGP32 to be present in the experiments.”

(21) Line 208 - Why were both magnitude and p-value criteria used rather than just p-values?

In the present proteomic approach each experimental condition was measured six times, and the average (mean) value was used to reduce random noise. Then we selected differences that had to be large enough to matter biologically, this is a central criterion and at least a 2-fold change was considered to focus exclusively on biologically relevant differences, which allowed us to control for the effect size. However, the differences also had to be statistically significant, we applied a statistical confidence at P < 0.01, to be sure that there is less than a 1% chance the result happened randomly. In the present proteomic approach each experimental condition was measured six times, and the average (mean) value was used to reduce random noise.

Then we selected differences that had to be large enough to matter biologically, this is a central criteria and at least a 2-fold change was considered to focus exclusively on biologically relevant differences, which allowed us to control for the effect size. However, the differences also had to be statistically significant, we applied a statistical confidence at P < 0.01, to be sure that there is less than a 1% chance the result happened randomly. We considered that using both criteria makes the results meaningful and trustworthy, not just a small or random fluctuation.

(22) Line 270 - Were these three replicate experiments also "independent"; if yes, in what sense?

“All experiments were conducted in triplicate” was replaced by “The experiments were performed using biological triplicates, each of which was analyzed in triplicate.”

(23) Line 296 - Perhaps "the temperature-sensitivity (or resistance) of" rather than "the nature of".

The modification was made in the new manuscript.

(24) Line 307 - The sentence mentions only one influential period that was removed from the dataset, but the word 'whenever' suggests multiple occurrences.

We agree, “whenever” was replaced by “because”.

(25) Line 325 - line 327 - The rationale behind the first part of the following sentence isn't clear to me, and what is meant by the second part is also not clear.

To clarify this point, “This result is consistent with the difficulty that Vibrio has in growing at temperatures below 20°C and with the complex interacting factors driving bloom dynamics (Laanaia et al., 2013)” was replaced by “This result is consistent with the difficulty Vibrio has in growing at temperatures below 20°C and with the many environmental factors that influence the dynamics of algae proliferation (Laanaia et al., 2013)."

(26) Line 327 - line 328 - Hard to interpret; does this refer to living algal cells, or all algal cells, living and degraded?

To improve clarity, “Interestingly, in spring 2015, the mean densities of all Alexandrium cells and of free-living Vibrio were positively correlated” was replaced by “Interestingly, in spring 2015, the mean densities of Alexandrium cells (living and degraded) and of free-living Vibrio were positively correlated”

(27) Figure 2 - These results strongly point to predation, but why the Vibrio population would already be elevated in the co-culture treatment relative to the control immediately after inoculation (0 hrs) is not clear.

The experiments were not conducted at the same time, and the first value on the graphs corresponds to the concentration of vibrio determined after 1 hour of exposure/incubation and not at time 0. Figures 2A and 2B have been modified accordingly, and substantial changes have been made to the relevant section of the results.

(28) Line 348 - There's no mention of Figure 2C in the main text, or of the statistical test associated with it in the Figure 2 legend.

To address this comment, Figure 2C has now been cited in the main text, and the statistical analysis method has been added to the Figure 2 caption.

(29) Line 352 - Text descriptions of videos are not easy to connect with the video content. Label the file names the same as how they are referred to in the text.

We agree with you, the sentence “Epifluorescence microscopy observation of GFP-labelled V. atlanticus LGP32 (previously grown in Zobell medium) in interaction showed that A. pacificum ACT03 cells that had lost their motility were attacked individually by V. atlanticus LGP32 before being lysed (Fig, 2C and Video 1). “was rephrased and replaced by “Epifluorescence microscopy observation of GFP-labelled V. atlanticus LGP32 (previously grow in Zobell medium) in interaction showed that V. atlanticus LGP32 simultaneously attacks A. pacificum ACT03 cells (Fig, 2C and Video 1).”

(30) Movie 1 could be cut to remove uninteresting footage at the start. What indicates lysis? Is the deformation of the cells an indication of lysis?

To respond to this comment, Video 1 has been shortened and in the caption, “degraded” was replaced by “lysed”

(31) Line 353 - Video could be zoomed in more on a few typical attacks to remove visual noise.

A chronological overview of an attack has been added to Figure 2 corresponding to Figure 2D, and a chronological overview of the overall event has been added to Figure 3 corresponding to Figure 3B1.

(32) Line 355 - There does not seem to be a Figure 3A2.

To address this point, the Fig. 2 and Fig. 3 has been revised for more clarity. See above

(33) Figure 3 - Can the authors fully exclude an effect of bacterial density as distinct from an effect of growth/starvation phase? It would be helpful to determine bacterial viable population densities at 12, 36, 60, and 126 hrs of incubation in Zobell medium, and to control for density in testing for effects on algae.

Information on Vibrio densities incubated in Zobell medium for 12, 36, 60, and 126 hours has been now included in the results section “Attack of A. pacificum ACT03 is activated by V. atlanticus LGP32 starvation.”

(34) Line 363 - It is unclear how the degradation of the flagella is apparent from movie 3. It would be helpful to have a comparison with healthy flagella.

Alexandrium cells with intact flagella move so quickly that it is impossible for us to follow them and film their flagella with the tools at our disposal.

For greater clarity, arrows have been added to videos 3, 4 and 5.

(35) Line 364 - Sudden change from referring to the recording as 'video' instead of movie. What is meant by erratic swimming? The cell does not seem to move much.

To address this comment, “Movie” was replaced by “Video” throughout the manuscript and “erratic swimming” was replaced by “irregular swimming”

(36) Line 365 - How did you observe the detachment of the flagellum?

The detachment of the flagellum can be observed using a confocal microscope. This process was filmed and presented in Video 3. Arrows have been added to the video to clearly indicate the flagellum detachment.

(37) Line 368 - Perhaps this is due to it not being clear regarding which movie is meant, but there is no clear attack visible in movie 4.

To make this clearer, arrows have been added to the video 4 to indicate attached cells.

And the sentence in the caption of the video 4 “Vibrio, filmed under a confocal microscope, attacks in groups one immobilized Alexandrium cell then moves on to attack — still as a group — another cell without touching the other whole cells, suggesting active communication between Vibrio cells” was rewritten and replaced by “This video, recorded under a confocal microscope, shows Vibrios simultaneously attacking a first immobilized Alexandrium cell, then moving on to attack a second cell without ever targeting the other cells present, suggesting active communication between the Vibrio bacteria.”

(38) Line 369 - It seems the peak attach % was reached at 45 minutes, not 15-30 minutes.

Sorry for the confusion. In fig. 3 for more clarity, the sentence “(A) Percentage of A. pacificum ACT03 motile cells. (B) cells attacked by V. atlanticus LGP32 and (C) cells lysis after 0, 15, 30, 45 and 60 min of interaction” was replaced by “(A) Cumulative percentage of motile A. pacificum ACT03 cells. (B) Cumulative number of cells attacked by V. atlanticus LGP32 and (C) Cumulative cell lysis after 0, 15, 30, 45 and 60 minutes of interaction.”

(39) Line 382 - "clearly show role of nutrient limitation", see comment re controlling for any role of bacterial density.

To address this point, information’s on Vibrio densities were added in the manuscript. See cf comment 33.

(40) Line 385 - line 386 - Phrasing unclear.

We have revised the text accordingly, “To this aim, A. pacificum ACT03 in exponential growth phase was first exposed for 30 min to supernatant from 60 hours starved V. atlanticus LGP32 Zobell media that induced 25% lysis of A. pacificum ACT03 cells and next to the corresponding V. atlanticus LGP32 cells. Group attacks were observed on non-degraded A. pacificum ACT03 cells, but not on lysed cells.“ was replaced by “To this end, A. pacificum ACT03 in exponential growth phase was first exposed for 30 minutes to the supernatant of a 126-hour culture of V. atlanticus LGP32, which induced lysis of 70% of the A. pacificum ACT03 cells (Figures 3C and 3C1, arrow 2 and video 4). Next, cells of V. atlanticus LGP32 from a 60-hour culture, capable of attacking A. pacificum ACT03 cells (Fig. 3B), were added. For 1 hour of exposure, no attack was observed on the previously lysed algae.”

(41) Line 413 - Is this the only pathway for quorum sensing in V. atlanticus?

Indeed, the last two sentences of this paragraph are unclear.

To address this point:

“By targeted mutagenesis of key genes involved in QS pathways ΔluxM (HAI-1 production), ΔluxS (AI-2 production) and ΔluxR (high-density QS master regulator) did not lead to any change in the attack behaviour of V. atlanticus LGP32 (Fig. 4C).” was replaced by “Targeted mutagenesis of key genes involved in two of the three known QS pathways in vibrios (Fig. S3), ΔluxM (HAI-1 production), ΔluxS (AI-2 production), and ΔluxR (main high-density QS regulator), did not result in any changes in the attack behavior of V. atlanticus LGP32 (Fig. 4C).”

And “Taken together these results showed that attack by V. atlanticus LGP32 is not link to QS.” was replaced by. “Combined with the absence of overexpression of the CqsS gene (inducible by CAI-1) involved in the last known QS pathway in Vibrio (Fig. S3), these results indicated that the attack by V. atlanticus LGP32 is most likely unrelated to QS.”

(42) The references to tropism aren't clear.

You're right, there's no reason to use the term tropism here. We have removed it.

(43) Line 439 - Why was H3BO4 used as a control for the addition of FeCl3?

For clarity, the sentence “Boron being known to be a regulator or capable of being transported by vibrioferrin (Romano et al., 2013; Weerasinghe et al., 2013), we tested its potential involvement in the interaction but no effect was evidenced here.” was replaced by “Given that boron is known for its role in regulating a global bacterial cellular response to phytoplankton and to bind to vibrioferrin (Romano et al., 2013; Weerasinghe et al., 2013), we tested its potential involvement in simultaneous vibrio attacks. Compared to the Zobell control, no effect on the number of attacks was observed”

(44) Line 441 - line 449 - Should explicitly say in text that no attacks were observed for any species other than the Alexandrium and Gymnodinium species.

We agree and have explicitly stated in the text that no attacks were observed for any species other than Alexandrium and Gymnodinium.

(45) Line 454 - line 455 - The last part of this sentence seems a strange statement, since

(i) it has long been know that predatory bacteria can eat a wide range of eukaryotes, ii) one of the cited papers (Perez et al) actually highlights a case of bacterial predation on algae, and iii) in the next paragraph the authors themselves highlight Streptomyces predation of algae.

To make this clearer, « Among predators, predatory bacteria are found in a wide variety of environments, and like bacteriophages and predatory protists, they have been reported to prey exclusively on other bacteria » was replaced by “Among predators, predatory bacteria are found in a wide variety of environments and, like bacteriophages and predatory protists, feed primarily on other bacteria, although a few cases of predation on microbial eukaryotes have also been reported.”

(46) Line 455 - Better to clarify the authors' definition of a predator at the start of the paper. The offered definition seems more like a definition of 'consumer' than 'predator', as the latter normally involves both the killing and consumption of other organisms, not just consumption with some kind of "expense".

To address this comment:

- “predator behaviour” was replaced by “predator-like behaviour”

- and “Considering predator as a free organism that feeds at the expense of another, this study is the first evidence of the capacity of some Vibrio to develop a predatory strategy against an alga. This behaviour differs from parasitism, because the survival of Vibrio is not exclusively dependent on algae in environment” was replaced by “Consider a predator as a free-living organism that kills its prey and feeds on it, this study provides data suggesting the ability of Vibrios to develop an original predator-like behaviour to kill and feed on algae.”

(47) Line 457 - Don't see the benefit of trying to distinguish from parasitism here, especially since parasitism can be facultative, whereas the authors' phrasing suggests that it is always obligate.

You are right, this sentence has been deleted.

(48) Line 463 - line 464 - The authors should clearly explain exactly what detailed aspects of Myxococcus and Lysobacter predation they think the "attack stage" of V. atlanticus resembles.

Accordingly, “The second stage, the ‘attack stage’ corresponding to physical contact between Vibrio and Alexandrium resembles the ‘wolf-pack attack’ strategy described for Myxococcus xanthus and Lysobacter regardless of the prey species used, M. xanthus must be in close proximity to prey cells in order to induce their lysis and to benefit from their biomass (Martin, 2002; Perez et al., 2014)” was replaced by “The second stage, the ‘attack stage’ corresponding to the physical contact between Vibrios and Alexandrium, is similar to the strategy used by Myxococcus xanthus and Lysobacter. These bacteria must be in close proximity to their prey in order to cause lysis and utilize their biomass, regardless of the prey's species (Martin, 2002; Genovesi et al., 2013; Perez et al., 2016; Zhang et al., 2020)”

(49) Line 466 - line 467 - The comparison to bacteria clustering around lysed cells is surprising since the authors show that V. atlanticus does not attack already lysed cells.

The sentence was rephrased, “This phenomenon is comparable to that of bacteria clustering around lysed ciliate cells “was replaced by “Visually, this phenomenon resembles bacteria clustering around lysed ciliate cells.”

(50) Line 469 - Missing is a statement of exactly what criteria constitute "wolf-pack hunting behaviour" and exactly how V. atlanticus meets those criteria.

To address this point, “wolf-pack hunting behaviour” was replaced by “predator-like behaviour”

'Able of' should be corrected to 'Capable of'.

We agree and have reworded the sentence.

(51) Line 470 - Consider starting a new paragraph for the material on quorum sensing.

Accordingly, we have separated the section concerning QS pathway from the section concerning iron pathway.

(52) As part of their discussion on the role of iron uptake, can the authors comment on any relationship between starvation and iron uptake, and in particular the observations that, while general nutrient deprivation induces attacks, supplementation with a specific nutrient (iron) also induces attacks (Figure 4D)? Do bacteria starved for general growth substrates take up more iron than growing bacteria?

To respond to this comment, “Future study could demonstrate further the role of vibrioferrin in group attack, by adding iron-saturated vibrioferrin to algae-Vibrio co-cultures.” was replaced by “Interestingly, if a general nutrient deficiency causes attacks, iron supplementation increases the number of attacks (Figure 4D), suggesting the importance of iron absorption in the attack behavior. Future studies should determine whether nutrient deficiency increases the iron absorption capacity of Vibrios and whether this plays a major role in the attack mechanism.”

(53) Line 486 - Of what is boron known to be a regulator?

To respond to this comment, “Given that boron is known for its regulatory properties and for being transportable by vibrioferrin“ was replaced by “Given that boron is known for its role in regulating a global bacterial cellular response to phytoplankton and to bind to vibrioferrin”.

Version 2

Abstract

Phytoplankton serve as a source of nutrients for bacteria in the marine environment. The interactions between algae and bacteria are known to include mutualism, commensalism, competition or antagonism. This occurs in the microenvironment surrounding phytoplankton cells, the phycosphere. An interface rich in nutrients and organic molecules exuded by the cell. Here, based on in situ observation and on an in vitro interaction study, we report on a novel form of starvation-induced hunting that the cells of selected Vibrio species exert on dinoflagellates. The results showed that Vibrio atlanticus was capable of attacking and killing the dinoflagellate Alexandrium pacificum ACT03. Briefly, the observed mechanism of algal-killing consists of first, the ‘immobilization stage’ involving the secretion of algicidal metabolites that disrupt the flagella of the algae. The ‘attack stage’ when Vibrios simultaneously surrounds algal cells at high density for a brief period without invading them. Finally, the ‘killing stage’ induces the lysis and consumption of the dinoflagellates.

By using a combination of biochemical, proteomic, molecular and fluorescence microscopy approaches, we showed that this relationship is not related to the decomposition of algal organic matter, Vibrio quorum sensing pathways, toxicity of the algae or pathogenicity of the bacterium but is conditioned by nutrient stress, iron availability and link to the iron-vibrioferrin transport system of V. atlanticus. This is the first evidence of a new mechanism that could to be involved in regulating Alexandrium spp. blooms and giving Vibrio a competitive advantage in obtaining nutrients from the environment. The interaction model we propose here suggests that Vibrio could play a role in regulating the proliferation of Alexandrium spp., giving it a competitive advantage in obtaining nutrients from the environment.

Introduction

Harmful algal blooms (HABs) have experienced an increase in their occurrence, intensity, and geographical distribution on a global scale, resulting in adverse environmental, health, and socioeconomic impacts (Marampouti et al., 2021). HABs have a considerable impact on human health as a result of direct exposure to volatile toxins or by toxic seafood consumption (Burkholder et al., 2018). From an ecological point of view, the expansion of HABs can result in the erosion of biodiversity, because they cause massive mortality of marine species and they are generally monospecific in nature (Chai et al., 2020). In coastal areas, understanding the biological interactions that control toxic algal blooms is therefore a major ecological challenge. Among HAB-causing organisms, a number of Alexandrium species have been placed on the list of invasive Mediterranean species. Among them, Alexandrium pacificum is a flagellated eukaryotic unicellular organism that together with Alexandrium tamarense and Alexandrium fundyense form the “Alexandrium tamarense” complex responsible for paralytic shellfish poisoning worldwide (Hadjadji et al., 2020). Since 1998, A. pacificum (former A. catenella) was monitored by the French phytoplankton observation and monitoring network (Rephy) in the Thau lagoon (French Mediterranean) because it produces paralytic shellfish toxins (PSTs) resulting in paralytic shellfish poisoning (PSP) syndrome.

Laania (Laanaia et al., 2013) showed that in Thau lagoon, a water temperature around 20°C for several days and organic and inorganic nutrients in sufficient concentrations are parameters favoring the development of A. pacificum, whose massive blooms occur in autumn. Algal blooms are seasonal events resulting in a rapid increase in the concentration of a species of algae in an aquatic environment.

Depending on the species of phytoplankton, tolerance to physicochemical parameters varies, which influences when blooms occur (Leblad et al., 2020). Interestingly, although the collapse of phytoplankton blooms has been previously attributed to viruses (Pal et al., 2020), some ecological studies have suggested an important role of algicidal bacteria (Su et al., 2007; Wang et al., 2010). Among them several are belonging to the Vibrio genus (Li et al., 2014; Wang et al., 2020).

Vibrio (class γ-proteobacteria) are common microorganisms in marine systems worldwide (Baker-Austin et al., 2017; Mavian et al., 2020), where they are important components of the food chain, particularly in biodegradation, nutrient regeneration and biogeochemical cycles (Oberbeckmann et al., 2012). Vibrio is one of the most studied bacterial taxa due to their ubiquity in coastal marine systems and their capacity to cause infections in humans and animals, leading sometimes to epizootic or zoonotic epidemics (LeRoux et al., 2015; Mavian et al., 2020). Vibrio are extremely adaptable to their environment (Johnson, 2013). The main factors influencing their occurrence and distribution in water are temperature, salinity, nutrient availability (Wang et al., 2020), multiple strategies such as biofilm formation on biotic and abiotic surfaces (Espinoza-Vergara et al., 2020), or interactions with a multitude of other organisms such as eukaryotic predators (Drebes Dörr and Blokesch, 2020) or plankton (Lopez-Joven et al., 2018) are used by Vibrio in the environment. There is also evidence that global climate change has increased Vibrio-associated illnesses affecting humans and animals (Brumfield et al., 2021; Muhling et al., 2017). However, the drivers and dynamics of Vibrio survival and propagation in the marine environment are not yet fully understood.

A substantial number of research articles have highlighted the potential of γ-proteobacteria to exert algicidal activity against dinoflagellates, supporting the hypothesis that γ-proteobacteria such as Vibrio play a role in the control of algae blooms in situ (Coyne et al., 2022). However, the mechanisms behind Vibrio-driven algal lysis in the environment remain to be elucidated. Particularly, it is unclear how in the water column, algicidal compounds secreted by bacteria can concentrate around the algae to exert their lytic effect.

This study aims to describe observations made in the natural environment between Vibrio bacteria and Alexandrium algal blooms, and to determine in vitro the main factors involved in this relationship. Using a combination of biochemical, proteomic, molecular, and fluorescence microscopy approaches, we explored the role of algal toxicity, bacterial pathogenicity and the quorum sensing pathway on this relationship and showed the important role of nutrient stress and the iron uptake pathway in this unique Vibrio/Alexandrium interaction.

Methods

Quantification of Alexandrium algae and Vibrio bacteria in the environment by Qpcr

Seawater samples were collected in the Thau Lagoon (southern France, a shallow Mediterranean ecosystem open to the sea (Abadie et al., 1999), during spring and autumn 2015. Briefly, samples were collected from the subsurface (-50 cm) near an oyster table at a phytoplankton surveillance site (part of the REPHY network, N 43°26.058’ and E 003°39.878’). Once a week during spring and autumn 2015, during field sampling campaigns, 20 L of water was filtered on board through a 180 μm pore-size nylon membrane. At the laboratory, according to Lopez-Joven et al. (Lopez-Joven et al., 2018) seawater was fractionated into two size classes as follows: 2 L of the above filtrate was filtered through a 0.8 μm pore-size polycarbonate Whatman Nuclepore membrane to obtain organisms in the 0.8–180 μm range corresponding to plankton-associated Vibrio and living Alexandrium forms. Then, the filtrate from the 0.8 µm membrane was filtered again through a 0.2 μm pore-size polycarbonate Whatman Nuclepore membrane until the membrane was saturated. Alexandrium cells, ranging from 25 to 40 µm, belong to the microphytoplankton and are therefore retained in the 0.8–180 µm fraction. Any Vibrio cells potentially associated with or attached to Alexandrium cells will also be retained in this fraction. Vibrio cells are approximately 0.5-0.8 µm-thick. The fraction between 0.2 and 0.8 µm therefore includes the free-living Vibrio. The bacterial population collected on 0.8-µm-pore-size filters was designated the particle-associated community, and the population on 0.2-µm-pore-size filters was designated as the free-living community. Membranes (in triplicate) were then conserved in 500 µL of 100% EtOH at -20°C. Environmental DNA (eDNA) was extracted from the MF Millipore membrane using the Macherey-Nagel NucleoSpin Tissue Kit and resuspended in 100 µL of water. The samples were then stored at -20°C after eDNA quantity and purity were assessed using a NanoDrop system (NanoDrop Technologies, Wilmington, DE, USA). PCR amplification reactions were done on a Roche LightCycler 480 Real-Time thermocycler (qPHD platform, University of Montpellier, France) using specific primer pairs (Table 3). Typically, the reactions contained 1 µL of template DNA (the DNA concentration for all samples varied from 1 to 40 μg mL−1), 0.5 µL of each primer (3.33 µM) and 4 µL of reaction mixture (SYBR Green Master Mix) in a total volume of 6 μL. The reaction parameters were as follows: 5 minutes at 95°C (initial denaturation) and 40 cycles of 10 s at 95°C (denaturation), 10 s at the corresponding hybridization temperature (Table 3) and 10 s at 72°C (elongation). Melting curve profiles were generated by increasing the temperature from 65°C to 95°C at 0.5°C s-1. Amplification products were analysed using LightCycler software (Roche Diagnostics). Vibrio spp. and A. pacificum and A. tamarense were quantified by constructing calibration curves based on DNA from the V. atlanticus LGP32 reference strain (former V. tasmaniensis LGP32) and from the A. pacificum reference strain (ACT03: A. catenella strain isolated form Thau in 2003) and the A. tamarense reference strain (ATT07: A. tamarense isolated from Thau in 2007) (not shown).

Strains and growth conditions

Vibrio strains

Wild-type and isogenic mutants of Vibrio atlanticus LGP32 (Table 1) were used in this study. Deletion-mutants included ΔluxR, ΔluxP, ΔluxM and ΔpvuB isogenic strains. The ΔpvuB mutant was constructed here by allelic exchange as described previously by Le Roux (Le Roux et al., 2007). We also used V. atlanticus LGP32 carrying the pSW3654T-GFP plasmid (Le Roux et al., 2007), hereafter referred to as V. atlanticus LGP32-GFP. Bacterial strains were grown at 22 ± 1°C in Zobell medium (0, 38 µM iron (III)). When needed, 25 μg mL-1 chloramphenicol (Cm) was added to cultures of V. atlanticus LGP32-GFP (Le Roux et al., 2007).

Phytoplancton strains

Non-axenic phytoplankton species (Table 2) were grown in batch culture in enriched natural sea water (ENSW, 6,55 µM iron (III)) with a salinity of 36 practical salinity units (PSU) at 22 ± 1°C under cool white fluorescent illumination (100 µmol photons m-2 s-1) and a 12 h:12 h light:dark cycle (Harrison et al., 1980). The algae were used for experiments in their exponential growth phase.

Co-culture assay

For each tested phytoplankton species (Table 2), 2.104 cells harvested in their exponential growth phase (doubling time between 5 and 7 days) were placed in 20 mL of ENSW medium in a 50 mL suspension culture flask (Cellstar® PS, Greiner bio-one). After incubation for 24 h at 22 ± 1°C under cool white fluorescent illumination, 40 µL of Vibrio strains (Table 1) grown for 12, 36, 60 and 156 h in Zobell medium or in Zobell medium supplemented with Fecl3 (6 µM iron (III)) or with H3BO3 (0.47 mM) or the corresponding culture supernatant were added to the phytoplankton cells (Table 2). After incubation at 22 ± 1°C for 0, 15, 30, 45 and 60 min under cool white fluorescent lights, living, non-swimming, attacked and lysed phytoplankton cells were counted in a sedimentation chamber under an inverted microscope. The number of lysed cells corresponded to phytoplankton cells showing disrupted membranes. Non-swimming algae were not counted as lysed cells. For Vibrio analysis, 100 µL of a 1:10 serial dilution mixture in ENSW (from 10-2 to 10-10) was plated on Vibrio Selective TCBS (thiosulfate-citrate-bile salts-sucrose) agar (in triplicate). After incubation for 24 h at 22 ± 1°C, the number of living Vibrio cells was determined by counting colony-forming units (CFUs). The data come from three independent experiments using independent phytoplankton cultures and independent bacterial cultures.

Microscope observations

The dynamic of the interaction between A. pacificum ACT03 isolated from the French Thau Lagoon, south of France (Laabir et al., 2011) and V. atlanticus LGP32, which is an oyster pathogen isolated from the French Atlantic coast (Gay et al., 2004) and present in the Thau Lagoon (Lopez-Joven et al., 2018) was surveyed. As the A. pacificum ACT03 strain (table 2) used in the study is not axenic, there is potential for bacteria other than V. atlanticus LGP32 to be present in the experiments. To elucidate the interaction without thoroughly accounting for the non-axenic cells, interaction was observed under a Zeiss Axio upright fluorescence microscope equipped with an AxioCamMRm 2 digital microscope camera using V. atlanticus LGP32 tagged with green fluorescent protein (GFP). Lasers were used at excitation wavelength (λex) 488 nm for GFP (emission wavelengths (λem): 505–530 nm) and λex 532 nm for plankton chloroplasts (λem 560–630 nm). Images were taken sequentially to avoid cross-contamination between fluorochromes. Sequences of images were merged during the Vibrio-Alexandrium interaction using ZEN 2012 (blue edition) software. Interaction events between Vibrio strains and phytoplankton strains were also observed under a Leica TCS SPE confocal laser scanning system connected to a Leica DM 2500 upright microscope camera (Montpellier RIO Imaging Platform, University of Montpellier, France).

Comparative proteomic analysis

Vibrio sampling and protein extraction

V. atlanticus LGP32 was grown for 60 h at 22°C in artificial seawater (high nutrient stress, cond. 1) or 12 h at 22°C in Zobell media (low nutrient stress, cond. 2). After 10 min centrifugation at 8000 rpm, crude protein extracts of V. atlanticus LGP32 in each culture condition (triplicates) were obtained by sonication on ice at 20% amplitude for 20 s in 200 µL of ice-cold denaturing buffer (7 M urea, 2 M thiourea, 4% CHAPS in 30 mM Tris-HCl, pH 8.5) and clarified by centrifugation at 2000 x g, 15 min, 4°C. The protein concentration of the supernatant was estimated using the 2D Quant Kit (Cytiva™, MERCK) and samples were stored at -80°C until use.

Bi-dimensional gel electrophoresis (2D gel)

Proteins extracts were individually analysed on 2D gel electrophoresis (6 gels per condition each corresponding to different biological replicates). To do so, 100 µg of proteins from each extract was added to rehydration buffer (7 M urea, 2 M thiourea, 4% CHAPS, 65 mM DTT) for a total volume of 350 µL. They were then individually loaded onto 17 cm isoelectric focusing strips (Bio-Rad) with a stabilized non-linear pH ranging from 3 to 10. Due to the high complexity of the protein profile in the acidic part (left) of the gel pH 3–10 (Fig. S3), we conducted additional ‘close-up’ analyses in gels using 17 cm isoelectric focusing strips (Bio-Rad) with a narrower, stabilized pH gradient ranging from 4 to 7. Strips were rehydrated passively for 5 h at 22°C, followed by active rehydration for 14 h under a 50 V current at 22°C (to help large proteins enter the strips).

Thereafter, isoelectrofocusing was carried out using the following programme: 50 V for 1 h, 250 V for 1 h, 8000 V for 1 h and a final step at 8000 V for a total of 140 000 V.h with a slow ramping voltage (quadratic increasing voltage) at each step. Focused proteins were reduced by incubating the strip twice in equilibration buffer (1.5 M Tris, 6 M urea, 2% SDS, 30% glycerol; bromophenol blue, pH 8.8) containing DTT (130 mM) at 55°C. Then, they were alkylated by incubation with equilibration buffer containing iodoacetamide (135 mM) on a rocking agitator (400 rpm) at room temperature protected from light.

Proteins were also separated according to their molecular weight (second dimension) on 12% acrylamide/0.32% piperazine diacrylamide gels run at 25 mA per gel for 30 min followed by 75 mA per gel for 8 h using a Protean II XL system (Bio-Rad). Gels were stained using an MS-compatible silver staining protocol and scanned using a ChemiDoc MP Imaging System (Bio-Rad) associated with Image Lab software version 4.0.1 (Bio-Rad).

Comparative bioinformatics analysis of 2D gels

Twelve gels (six per condition) were selected for comparative analysis on PD-Quest v. 7.4.0 (Bio-Rad) to identify changes in protein abundance between the proteomic profiles of V. atlanticus LGP32 cultured in contrasting nutrient conditions (ENSW/Zobell). Spots whose mean intensity across six replicates per strain was two times higher or lower than those from the other strain, with a P <0.01 (Mann-Whitney U-test), were considered significantly different in terms of abundance between the two conditions (quantitative difference). Differentially represented spots were then excised from the gels, destained, trypsin-digested and the obtained peptides were identified by tandem mass spectrometry (MS-MS) using the PISSARO platform facility (University of Rouen, France). To identify protein(s) present in each spot, the obtained peptides were compared with V. atlanticus LGP32 reference genome (https://vibrio.biocyc.org/). The genes whose peptides matched strongly were retrieved and used for an BLASTx query against non-redundant databases to determine the protein identity of the best match. A gene was considered as strongly matched when at least two peptides matched the sequence with a coverage of> 6%. Their theoretical isoelectric point (pI) and molecular weights were also calculated using the Expasy server (https://www.expasy.org/) to compare them with the location of the spot on the gel. Altogether, these complementary analyses made it possible to characterize the protein identity of each spot with confidence.

Gene expression analysis

Vibrio sampling and RNA extraction

V. atlanticus LGP32 was grown in Zobell media for 36 h and 60 h at 22°C (decline phase of growth, nutrient stress) or for 12 h at 22°C (exponential grow phase, poor nutrient stress). Total RNA was isolated from V. atlanticus LGP32 using the standard TRIzol method (Invitrogen Life Technologies SAS, Saint-Aubin, France) and then treated with DNase (Invitrogen) to eliminate genomic DNA contamination. After sodium acetate precipitation, the quantity and quality of the total RNA were determined using a NanoDrop spectrophotometer and agarose gel electrophoresis. Following heat denaturing (70°C for 5 minutes), reverse transcription was performed using 1 μg of RNA prepared with 50 ng μL-1 oligo-(dT) 12–18 mer in a 20-μL reaction containing 1 mM dNTPs, 1 unit μL-1 RNAseOUT and 200 units μL-1 Moloney murine leukaemia virus reverse transcriptase (M-MLV RT) in reverse transcriptase buffer, according to the manufacturer’s instructions (Invitrogen Life Technologies SAS, Saint-Aubin, France).

PCR amplification

Amplification reactions were analysed using a Roche LightCycler 480 Real-Time thermocycler (Bio-Environnement platform, University of Perpignan, France). In this study, several PCR primer pairs were designed using Primer3 software (optimal primer size: 20 bases; Tm: 60°C; primer GC%: 50; 2GC clamp and product size range: 150–200 bp) and calibrated with V. atlanticus LGP32 genomic DNA (Table 1). To determine the qPCR efficiency of each primer pair used, standard curves were generated using seven serial dilutions of genomic DNA (10-2, 10-3, 10-4, 10-5, 10-5, 10-7 and 10-8) (not shown); the qPCR efficiencies of the tested genes varied between 1.85 and 2.08 (Table 3). For gene expression, reverse transcription was performed with 1 µg of total RNA using random hexamers and SuperScript IV reverse transcriptase (Invitrogen). The total qPCR reaction volume was 10 μL and consisted of 5 μL of cDNA (diluted 1:5), 2.5 μL of SensiFAST SYBR No-ROX Mix (Bioline) and 100 nM or 300 nM PCR primer pair (Table 3). The reaction parameters were as follows: 2 min at 95°C (initial denaturation) and 40 cycles of 5 s at 95°C (denaturation), 10 s at 59°C (annealing) and 20 s at 72°C (elongation). The specificity of each PCR was checked by measuring fluorescent signals during melting curve analysis (PCR product heated from 65°C to 95°C continuously and slowly at 0.1°C s−1). Relative expression was calculated by normalization to the expression of two constitutively expressed housekeeping genes, namely, 6PKF (VS_2913) and CcmC (VS_0852), using the delta-delta threshold cycle (ΔΔCt) method(Pfaffl, 2001).

Detection of quorum-sensing signalling molecules

Vibrio culture

To detect the QS molecules (AI-2, AI-1 and CAI-1) V. atlanticus LGP32 was grown in Zobell media for 12 h (exponential growth phase, control) and 60 h (decline phase of growth, nutrient stress).

AI-2 analysis

Bioluminescence assay using the QS bioluminescent of Vibrio campbellii MM32 (luxN::Cm, luxS::Tn5Kan) was used to detect AI-2 molecules in culture supernatants. Briefly, Vibrio cultures were centrifuged at 17,000 x g for 10 min, and the resulting supernatants were filtered on 0.22 μm. Then 20 μL of the filtrates were mixed with 180 μL of V. campbellii MM32 diluted 1:5000 then incubated at 30°C and 100 rpm. Luminescence and cell density (OD620) were collected in triplicate and analysed according to Tourneroche et al.(Tourneroche et al., 2019).

AI-1 and CAI-1 extraction and LC-MS analysis

Chemical analyses were conducted with a Q Exactive Focus Orbitrap System coupled to an Ultimate 3000™ ultrahigh-performance liquid chromatography (UHPLC) system (Thermo Fisher Scientific) according to Rodrigues et al. (Rodrigues et al., 2022).

Briefly, ethyl acetate (2 mL) was added into each culture (2 mL). This mixture was shaken overnight at room temperature (150 rpm). The two phases were then separated and the aqueous phase was extracted once again. The two obtained organic phases were pooled and the solvent was evaporated under vacuum. The crude extracts were dissolved in 500 µL LC-MS grade methanol for analysis. The experiments were performed using biological triplicates, each of which was analyzed in triplicate. Analyses of extracts and standards (3 μL injected) were performed in electrospray positive ionization mode in the 50–750 m/z range in centroid mode. The parameters were as follows: spray voltage: 3 kV; sheath flow rate: 75; aux gas pressure: 20; capillary temperature: 350°C; heater temperature: 430°C. The analysis was conducted in Full MS data-dependent MS2 mode (Discovery mode). Resolution was set to 70,000 in Full MS mode, and the AGC (automatic gain control) target was set to 1x106. In MS2, resolution was 17,500, AGC target was set to 2x105, isolation window was 0.4 m/z, and normalized collision energy was stepped to 15, 30 and 40 eV. The UHPLC column was a Phenomenex Luna Omega Polar C18 1.6 μm, 150 x 2.1 mm. The column temperature was set to 42°C, and the flow rate was 0.4 mL min-1. The solvent system was a mixture of water (A) with increasing proportions of acetonitrile (B), with both solvents modified with 0.1% formic acid. The gradient was as follows: 1% B 3 min before injection, then from 1 to 15 min, a gradient increase of B up to 100% (curve 5), followed by 100% B for 5 min. The flow was injected into the mass spectrometer starting immediately after injection. All data were acquired and processed using FreeStyle 1.5 software (Thermo Fisher Scientific).

Chemicals and solvents

N-acyl-homoserine lactones (AHL) were obtained from Cayman Chemical (Ann Arbor, MI, USA). Stock solutions were obtained by dissolving standards in methanol or dichloromethane (C18-AHL) at a concentration of 1 mg mL-1 and stored at -80°C. Standard solutions for UHPLC-high-resolution tandem mass spectrometry (HRMS) analyses were prepared by diluting each individual standard solution with methanol in order to obtain a concentration range from 2000 to 20 ng mL-1. LC-MS grade methanol, acetonitrile and formic acid were purchased from Biosolve (Biosolve Chimie, Dieuze, France), analytical-grade ethyl acetate was obtained from Sigma-Aldrich. Pure water was obtained from Elga Purelab Flex System (Veolia LabWater STI, Antony, France).

Nature of lytic compounds secreted by V. atlanticus LGP32

To determine the temperature-sensitivity of the lytic compounds secreted, V. atlanticus LGP32 grown for 60 h in Zobell media at 22 °C was filtered through a 10 kDa membrane (Amicon® Ultra-4 filter unit). The eluate containing molecules with MW below 10 KDa was then incubated in a water bath at 100°C for 30 min. Boiled filtrates (0.1 mL) were subsequently used to inoculate A. pacificum (ACT03 strain) cultures, then lytic activity was observed under the Leica TCS SPE confocal laser scanning system. Zobell media with the same treatment was used as control.

Statistical analysis

Environmental data

Statistical analyses were performed using R 3.6.3 software. The relationship between Alexandrium and Vibrio was explored separately in spring and autumn. We used a generalized linear model specifying a Gaussian family. For spring, the dataset for salinity and temperature was complete (14 periods of observation). An influential period was detected and removed from the dataset because no dead Alexandrium cells were observed. The effects of explanatory variables such as log10 (Vibrio+1), salinity and temperature were centred, reduced and tested as fixed effects with a linear relationship. Model selection was performed using the Akaike information criterion corrected for small sample size (AICc). Models were considered different whenever the difference between their AICc value and the lowest AICc value (ΔAICc) was lower than 2 (Burnham and Anderson, 2002). Alexandrium distribution and model residuals were checked for normal distribution assumptions (QQ plot and Shapiro-Wilk test). For autumn, the dataset was complete for 10 periods of observation. Salinity and temperature were missing for three periods. We explored the relationship between Alexandrium and Vibrio alone using the method detailed above for spring.

In vitro data

Statistical analyses were performed using one-way ANOVA (analysed by pair) followed by Tukey’s test (Statistica 10.0 software, StatSoft, Maison-Alfort, France). *P <0.05, **P <0.01, ***P <0.001

Results

Concomitant occurrence of A. pacificum ACT03, A. tamarense ATT07 and free-living Vibrio spp. in the Thau lagoon

In the spring and autumn of 2015, in the Thau Lagoon (Fig. 1A), we detected Alexandrium algae (A. pacificum ACT03 and A. tamarense ATT07, both alive and in degraded cell forms) and free-living Vibrio, but no plankton-associated Vibrio were observed (Fig. 1B, Data S1). Using model selection based on AICc, we found no significant relationship between Alexandrium (A. pacificum ACT03, A. tamarense) and Vibrio Spp. abundances in autumn. This result is consistent with the difficulty Vibrio has in growing at temperatures below 20°C and with the many environmental factors that can influence the dynamics of algae proliferation (Laanaia et al., 2013). Interestingly, in spring 2015, the mean densities of all Alexandrium cells (degraded and alive) and of free-living Vibrio were positively correlated. The lowest AICc was obtained with the model explaining degraded form of Alexandrium density based on the free Vibrio density (Fig. 1C). Given that, this model is not so different from the model with only the intercept, but better than any other linear combination with other potentially interfering drivers, such as temperature and salinity (Fig. 1D), we searched for evidence of a relationship between Vibrio and Alexandrium by studying their interaction in vitro.

Dynamics of Alexandrium and Vibrio in the environment.

(A) Location of the monitoring station in the Thau Lagoon (southern France). (B) Mean abundance (DNA equiv.) of Vibrio spp. (16S) and Alexandrium spp. (A. pacificum ACT03 + A. tamarense ATT07). Vibrio cells (black line with diamond dot) and degraded Alexandrium cells (grey line with round dot) were evidence in the 0.2–0.8 μm fraction (free Vibrio fraction) in spring and autumn 2015. Living Alexandrium cells (grey line with roundot) but no plankton-associated Vibrio spp. (black line with diamond dot) were evidence in the 0.8–180 μm in spring and autumn. (C) Result of Akaike information criterion (AICc) models tested to explain the mean value of degraded Alexandrium cells (dead cells) in spring. (D) Wald test of the AICc model attributing the mean value of degraded cells of Alexandrium in spring to free Vibrio.

Vibrio atlanticus LGP32 feeds on Alexandrium pacificum ACT03

To investigate whether Alexandrium interacts with Vibrio, we incubated in Enriched Natural SeaWater (ENSW) A. pacificum ACT03 (2.0 x 104 cells) with V. atlanticus LGP32 previously grown for 12 hours in Zobell media (initial concentration of 8.8 x 107 cells mL-1). In interaction with V. atlanticus LGP32, A. pacificum ACT03 cell abundance decreased significantly from 2.1 104 cells mL-1 after 1h of exposure to 1.1 x 104 cells mL-1 after 48 h of exposure (Fig. 2A), while the V. atlanticus LGP32 concentration grew significantly after 26 h of interaction, reaching a maximum peak density of 7.6 x 107 CFU mL-1 at 34 h (Fig. 2B). In the control experiment where A. pacificum ACT03 was cultured alone in ENSW, the algal concentration remained stable over time (Fig. 2A) and no bacteria were on the corresponding TCBS plates (Vibrio selective medium). In the control where V. atlanticus LGP32 was grown alone in ENSW, the bacterial concentration decreased from 7.0 × 10⁵ CFU/mL after 1 hour of incubation to 1.1 × 10⁵ CFU/mL after 48 hours of incubation (Fig. 2B). These results show that the interaction between V. atlanticus LGP32 and A. pacificum ACT03 leads to a decline in the algal population and promotes the growth of V. atlanticus LGP32. This suggests that V. atlanticus LGP32 is able to feed on A. pacificum ACT03.

Incubation of Vibrio atlanticus LGP32 and Alexandrium pacificum ACT03 in enriched natural seawater (ENSW).

(A) A. pacificum ACT03 cultured alone (grey bar) and incubated with V. atlanticus LGP32 (black bar) in ENSW. (B) V. atlanticus LGP32 cultured alone (grey bar) and incubated with A. pacificum ACT03 (black bar) in ENSW.(C) Snapshot of the interaction between V. atlanticus LGP32-GFP cells (60-hour culture) and one cell of A. pacificum ACT03 taken at 8h00 of co-culture. (D) Chronological snapshots of the interaction (98 pictures, one per second). V. atlanticus LGP32 (small green cells) and A. pacificum ACT03 cell (large red cell). All experiments were done in triplicates. Asterisks indicate significant differences in a multiple comparison test (One-way ANOVA with post hoc Tukey test), *P ≤ 0.05.

V. atlanticus LGP32 performs attacks on A. pacificum ACT03

Epifluorescence microscopy observation of GFP-labelled V. atlanticus LGP32 (previously grow in Zobell medium) in interaction showed that V. atlanticus LGP32 cells are capable to simultaneously attacks A. pacificum ACT03 cells (Fig, 2C and Video 1). The attacks were extremely rapid, with empty thecae (algal envelopes) observed in the medium after less than 60 s (Fig 2D and Video 2). During the attack, V. atlanticus LGP32 did not invade the algal cell but remained clustered on the cell surface (Fig. 2C).

Attack of A. pacificum ACT03 is activated by V. atlanticus LGP32 starvation

In 2002, Martin hypothesized that nutritional stress induces bacteria to lyse algae (Martin, 2002). To test this hypothesis, we monitored V. atlanticus LGP32 behaviour in response to starvation (Fig. 3). We observed that V. atlanticus LGP32 in exponentially growth phase (12 h of culture in Zobell medium, 8.8 x 107 cells mL-1) did not interact with A. pacificum ACT03 cells for the first hour of contact (Fig. 3A, B, C). In contrast, V. atlanticus LGP32 in the decline phase (36 h of culture in Zobell medium, 1.3 107 cells mL-1) induced a significant decrease in the number of motile algae cells by 8.9% after 15 min and by 43.3% after 60 min (Fig. 3A). This phenomenon corresponded to the degradation and/or disruption of algal flagella (Video 3). The flagella no longer functioned correctly, which caused irregular swimming of the algae (Video 3, left cell). This was followed by a complete cessation of swimming. When the flagellum detached from the algae (Video 3, right cell), the attack occurred. With starved V. atlanticus LGP32 (60 h of culture in Zobell medium, 0.6 107 cells mL-1), algae immobilization was fast and significant (91.4% in 15 min, Fig. 3A), and algae were attacked individually being targeted by V. atlanticus LGP32 cells (Video 4). The percentage of cells attacked and killed peaked at 30% after 15–30 min of contact (Fig. 3B and 3B1) and then decreased. After 1 h, attacks had stopped with approximately 40% of the algal cells still alive (Fig. 3B1 and 3C). Although it remains unclear whether the attacks occur during a specific phase of growth, it is evident that the cells are already weakened before attack as they have all lost their flagella.

Role of Vibrio atlanticus LGP32 starvation in the interspecific interaction process.

Experiments were conducted by incubatingA. pacificum ACT03 with V. atlanticus LGP32 previously grown for 12, 36, 60 and 126 h in Zobell medium. (A) Cumulative percentage of motile A. pacificum ACT03 cells. (B) Cumulative number of cells attacked by V. atlanticus LGP32 and (C) Cumulative cell lysis after 0, 15, 30, 45 and 60 minutes of interaction. Corresponding pictures showing (A1) Black arrows indicate unhooked and degrade flagellum from A. pacificum ACT03 flagellum, (B1) Chronological sequence of five snapshots showing V. atlanticus LGP32-GFP cells (60-hour culture) and A. pacificum ACT03 cells, during the first hour of their interaction. V. atlanticus LGP32 (small green cells), living A. pacificum ACT03 (large red cells) and dead A. pacificum ACT03 (large green cell). (C1) Black arrow 1 indicate vesicle formation on A. pacificum ACT03 cell and black arrow 2 indicate lysed A. pacificum ACT03 cell. All percentages were determined based on a minimum of 2,000 cells of A. pacificum ACT03. All experiments were done in triplicates. Asterisks indicate significant differences in a multiple comparison test (One-way ANOVA with post hoc Tukey test), ***P ≤ 0.001

An old-starved culture of V. atlanticus LGP32 (126 h of culture in Zobell medium, < 0.1 107 cells mL-1) significantly immobilized A. pacificum ACT03 cells within a few minutes, with lysis occurring immediately (Fig. 3A and 3C), making it impossible to detect attacks by V. atlanticus LGP32 (Fig. 3B). The lysis phase corresponded to initial vesicle formation followed by the bursting of A. pacificum ACT03 cells (Fig. 3C and 3C1).

We next tested whether this lytic effect was linked to Vibrio culture supernatant and mediated by thermostable molecule (s) secreted by Vibrio. The culture supernatant of starved culture of V. atlanticus LGP32 (36 h) filtered through a 10 kDa membrane and then incubated at 100°C for 30 min still possessed its lytic properties, indicating that the algicidal compounds produced by V. atlanticus LGP32 are small thermostable molecules unlikely to be lytic enzymes, or lysins able to digest the algae cell. However, these experimental observations clearly show the key role of nutrient limitation in triggering the attack behaviour and the secretion of lytic compounds of V. atlanticus LGP32.

Attack occurs on A. pacificum ACT03 in exponential phase of growth

Here, we wondered whether the live/dead status of algae is important for V. atlanticus LGP32-mediated attacks targeting. To this end, A. pacificum ACT03 in exponential growth phase was first exposed for 30 minutes to the supernatant of a 126-hour culture of V. atlanticus LGP32, which induced lysis of 70% of the A. pacificum ACT03 cells (Figures 3C and 3C1 (arrow 2) and Video 4). Next, cells of V. atlanticus LGP32 from a 60-hour culture, capable of attacking A. pacificum ACT03 cells (Fig. 3B), were added. For 1 hour of exposure, no attack was observed on the previously lysed algae. This result is similar to what is observed on the Video 1 with flash attacks only on immobilized, but not degraded A. pacificum ACT03 cells (red), and not on lysed cells (green). In addition, no attacks occurred on cells from an old A. pacificum ACT03 culture (1-month culture). Together, with the very short duration of attacks (Video 1), these results indicate that V. atlanticus LGP32 attacked exponentially growing cells of A. pacificum ACT03, but not decomposing algae, suggesting that this behaviour is not just an opportunistic response of heterotrophic bacteria to organic substrates.

Attack is independent of quorum sensing

Considering the simultaneous action of V. atlanticus LGP32 in attacks on A. pacificum ACT03, we tested whether the attack process depended on the key physiological mechanism that regulates many functions in marine microbial cells, quorum sensing (QS) (Lami, 2019; Papenfort and Bassler, 2016), a type of cell-cell communication. Although QS is a cell-density-dependent mechanism, our results showed no attack from a 12 h culture of V. atlanticus LGP32 up to a concentration of 4.106 Vibrio mL-1 (Fig 4A). Attacks were only observed with V. atlanticus LGP32 from a 60 h culture at low concentration of 5.103 Vibrio mL-1 to the highest concentration tested of 5.105 Vibrio mL-1 (Fig 4A), consistent with the hypothesis that the attacks were independent of Vibrio density.

Role of quorum sensing and the vibrioferrin iron uptake pathway in the interaction process.

(A) Effect of V. atlanticus LGP32 cell density on the attack process. A. pacificum ACT03 cells (1.103 cells mL-1) were incubated with V. atlanticus LGP32 grown for 60 hours in Zobell medium at concentrations ranging from 5.103 to 5.105 cells mL−1 (black bars). For comparison, A. pacificum ACT03 incubated with V. atlanticus LGP32 grown for 12 hours in Zobell medium at concentrations ranging from 5.103 to 4.106 cells mL−1 (grey bars). The image on the bars indicates either unaffected algae (live red algae) or algae attacked by vibrio’s (algae covered with green vibrio’s) during the interaction (B) CqsS, luxM, luxN, luxS, and luxP quorum sensing and PvsA, PvuB and PvuA2 vibrioferrin pathway genes expression in V. atlanticus LGP32 grown for 12, 36 and 60 h in Zobell medium. (C) Effect of V. atlanticus LGP32 mutants on the attacked process. Experiments were conducted by incubating A. pacificum ACT03 with V. atlanticus LGP32, V. atlanticus LGP32 tagged with GFP, V. atlanticus LGP32 washed with ENSW or V. atlanticus LGP32 mutant ΔPvuB, ΔluxM, ΔluxR and ΔluxS previously grown 60 h in Zobell media (control), The percentage of A. pacificum ACT03 attacked was determined during the first 30 min of exposure. (D) Effect of V. atlanticus LGP32 cultures media composition on the attacked process. Experiments were conducted by incubating A. pacificum ACT03 with V. atlanticus LGP32 grown 60 h in Zobell media supplemented with H3BO4 or FeCl3. The results were compared with an exposure to V. atlanticus LGP32 grown 60 h in Zobell media. All percentages were determined based on a minimum of 2,000 cells of A. pacificum ACT03. All experiments were done in triplicates. Asterisks indicate significant differences in a multiple comparison test (One-way ANOVA with post hoc Tukey test), **P ≤ 0.01, ***P ≤ 0.001

The analysis of the expression of genes involved in the known QS pathways in Vibrio cell (Fig. S3A), highlighted that only the AI-2 pathway was induced during nutrient stress of V. atlanticus LGP32, because only the expression of the AI synthase (LuxS) and its receptor (LuxP) increased significantly (Fig. 4B; ANOVA p <0.05). This was confirmed by a QS bioluminescence assay, which showed a AI-2 molecules (unquantified) in the Zobell culture supernatant of V. atlanticus LGP32 after 60 h of culture but not after 12 h of culture and not in the ENSW supernatant of V. atlanticus LGP32 after 12 or 60 h of culture.

UHPLC-HRMS/MS provided no evidence of detectable HAI-1) and CAI-1 in any experiments.

Targeted mutagenesis of key genes involved in two of the three known QS pathways in vibrios (Fig. S3), ΔluxM (HAI-1 production), ΔluxS (AI-2 production), and ΔluxR (main high-density QS regulator), did not result in any changes in the attack behavior of V. atlanticus LGP32 (Fig. 4C). Combined with the absence of overexpression of the CqsS gene (inducible by CAI-1) involved in the last known QS pathway in Vibrio (Fig. S3), these results indicated that the attack by V. atlanticus LGP32 is most likely unrelated to QS.

Attack related to the availability of iron

The comparative analysis of the proteome of V. atlanticus LGP32 incubated 60 h in artificial seawater (ENSW) versus V. atlanticus LGP32 grown 12 h in Zobell nutrient-rich medium revealed 10 proteins modulated by nutrient stress (Fig. S2). The two most down-regulated proteins correspond to be ß-ketoacyl-(acyl-carrier-protein) synthase II (-22-fold in ENSW compared to Zobell), a key regulator of bacterial fatty acid synthesis, and the dihydroorotase (-6.6-fold in ENSW compared to Zobell), an enzyme essential for pyrimidine biosynthesis and thus bacterial proliferation and growth. The low expression of these proteins in ENSW is consistent with V. atlanticus LGP32 nutritional starvation. The most up-regulated protein in starved V. atlanticus LGP32, with an increase of more than 6-fold, was glucosamine-6-phophate deaminase, an enzyme involved in bacterial energy metabolism probably necessary for its survival. Among the other up-regulated proteins, one was an iron siderophore-binding protein (Spot 4413, Fig. S2A) corresponding to the vibrioferrin outer membrane receptor PvuB, whose gene is part of the pvu operons involved in iron transport (Fig. S3B). Interestingly, the corresponding gene pvuB as well as the vibrioferrin membrane receptor gene pvuA2 (Fig. S3B) were both significantly induced in Vibrio under nutrient stress (Fig. 4B; ANOVA p <0.01) but not the one involved in the vibrioferrin biosynthesis, pvsA (Fig. 4B). Remarkably, among the 10 proteins identified by proteomic analysis only V. atlanticus LGP32 mutant lacking pvuB failed to attack A. pacificum ACT03 (Fig. 4C; ANOVA p <0.001). In the absence of the pvuB gene, V. atlanticus LGP32 was unable to attack simultaneously A. pacificum ACT03. In addition, V. atlanticus LGP32 cells that had been washed with ENSW to remove their culture supernatant metabolites also failed to attack A. pacificum ACT03 (Fig. 4C; ANOVA p <0.001), which is congruent with the hypothesis that attacks depend on the V. atlanticus LGP32 vibrioferrin transport system. Finally, attacks increased significantly, when FeCl3 was added to the Vibrio culture medium (Fig. 4D) but not with H3BO4 (Fig. 4D) Finally, attacks increased significantly, when FeCl3 was added to the Vibrio culture medium but not with H₃BO₄ known also to be capable of being transported by vibrioferrin (see Fig. 4D).. Taken together, those results are consistent with the hypothesis that attacks are regulated by iron.

Attack is a Vibrio spp. behaviour specific to Alexandrium spp

To evaluate the dinoflagellates specificity of the attack behaviour, a selection of Vibrio spp. was co-cultured with a selection of dinoflagellate strains commonly found in the Mediterranean Sea. The results showed that, among the Vibrio spp. tested (pathogenic or not) all, under nutrient stress, were able to secrete algicidal compounds, immobilize, attack and lyse A. pacificum ACT03 cells (Table 1) and no linked to their pathogenesis for fish of invertebrates was observed (Table 1). Among the sixteen dinoflagellates species tested, only Alexandrium spp. (non-toxic and paralytic shellfish toxin (PST) producers) and Gymnodinium catenatum (PST producer) were immobilized, attacked and lysed by V. atlanticus LGP32 (Table 2), but no linked to PSTs was revealed (Table 2).

Discussion

Predation is a widespread mode of interaction for survival in the natural world (Finke and Denno, 2004; Sinclair et al., 2003). Among predators, predatory bacteria are found in a wide variety of environments and, like bacteriophages and predatory protists, feed primarily on other bacteria, although a few cases of predation on microbial eukaryotes have also been reported (Johnke et al., 2014; Perez et al., 2016).

Consider a predator as a free-living organism that kills its prey and feeds on it, this study provides data suggesting the ability of Vibrios to develop an original predator-like behaviour to kill and feed on algae.

In fact, the strategy developed by Vibrio to kill algae may be reminiscent of strategies previously described in the prokaryotes (Johnke et al., 2014). As shown in Video 1, the interaction between V. atlanticus LGP32 and A. pacificum ACT03 proceeds in three stages (Fig. 5). The first stage, the ‘immobilization stage’, recalls the strategy used by Streptomyces to immobilize its prey (Kumbhar et al., 2014) based on the secretion of algicidal metabolites that disrupt the flagella. The second stage, the ‘attack stage’ corresponding to the physical contact between Vibrios and Alexandrium, is similar to the strategy used by Myxococcus xanthus and Lysobacter. These bacteria must be in close proximity to their prey in order to cause lysis and utilize their biomass, regardless of the prey’s species (Martin, 2002; Genovesi et al., 2013; Perez et al., 2016; Zhang et al., 2020). V. atlanticus LGP32 also surrounds A. pacificum ACT03 cells at high density for a very short time, but does not invade the algal cell. Visually, this phenomenon resembles bacteria clustering around lysed ciliate cells (Blackburn et al., 1998). The third stage, the ‘killing stage’, is similar to that of epibiotic bacterial predators, which induces the lysis of bacteria from the outside (Rashidan and Bird, 2001). Overall, these observations suggest that V. atlanticus LGP32 can exhibit a predator-like behaviour.

Schematic representation of a putative strategy developed by Vibrio spp. to feed on Alexandrium spp. and G. catenatum in the environment.

(A) Vibrio in the environment when subjected to starvation secrete non-protein lytic compounds. (B) Some of these lytic compounds degrade the flagella, immobilizing the alga (immobilization stage). (C) Then Vibrio swims and clusters around its prey (attack stage). (D) Lytic compounds released by Vibrio where able to concentrate around the algae cells, thereby lysing the algae (killing stage). (E) Feeding on the released nutrients, Vibrio multiply and then spread in the environment. Yellow clouds: Lytic compound release by Vibrio, Grey clouds: Algal nutrients released upon lysis.

The attack behaviour of V. atlanticus LGP32 was linked to iron absorption, as mutants with impaired iron absorption completely lost its ability to attack the algae. Iron is an essential element for growth in most organisms, including phytoplankton (Martin and Fitzwater, 1988) and bacteria (Neilands, 1981), and its concentration in seawater is known to be very low with measurements in the open ocean surface at 0.2 nM and in deep waters at 0.6 nM (Millero, 1998).

Moreover, Its low solubility in seawater limits its availability (Bruland et al., 1994; Wu and Luther, 1994). To acquire iron, bacteria have developed systems based on the secretion (and subsequent uptake) of iron-chelating siderophores to obtain this element from the environment (Amin et al., 2009a). Therefore, many Vibrio spp. produce a siderophore known as vibrioferrin, which is synthesized and secreted by proteins encoded by the pvsABCDE gene cluster (Fig. S3). Given that boron is known for its role in regulating a global bacterial cellular response to phytoplankton and to bind to vibrioferrin (Romano et al., 2013; Weerasinghe et al., 2013), we tested its potential involvement in simultaneous vibrio attacks. Compared to the Zobell control, no effect on the number of attacks was observed. For iron-vibrioferrin uptake, Vibrio parahaemolyticus uses a membrane siderophore receptor, called PvuA, which is coupled to an inner membrane ATP-binding cassette (ABC). This ABC transporter system comprised of proteins encoded by the pvuABCDE gene cluster (Fig. S3) is required for transporting the siderophore across the inner membrane (Tanabe et al., 2003). Siderophores are not only iron carriers but also important regulators of virulence (Miethke and Marahiel, 2007) and mediators of bacterial interaction with phytoplankton (Amin et al., 2009b; Kramer et al., 2020). We showed here a pivotal role of iron in the interaction between V. altlanticus LGP32 and A. pacificum ACT03. This mirrors the mutualistic interaction observed between Gymnodinium catenatum and Marinobacter (Amin et al., 2009b). In fact, in natural settings, the co-occurence of Marinobacter and G. catenatum is suggested to depend on a mutually beneficial utilization of iron and carbon resources (Bolch et al., 2011). As in the present study, iron seems to play a key role in the interaction. Indeed, the labile iron released through the photolysis of ferric chelates with vibrioferrin providing a crucial iron source for phytoplankton, which need substantial amounts of iron to support carbon fixation through photosynthesis (Amin et al., 2009a; Yang et al., 2021). This fixed carbon, in turn, sustains the growth of both the phytoplankton and their bacterial counterparts (Amin et al., 2009b; Kramer et al., 2020). Interestingly, if a general nutrient deficiency causes attacks, iron supplementation increases this number of attacks (Figure 4D), suggesting the importance of iron absorption in the attack behavior.

Future studies should determine whether nutrient deficiency increases the iron absorption capacity of Vibrio bacteria and whether this could plays a major role in the attack mechanism.

This study showed that quorum sensing is not involved in microbial attacks. Thus, Vibrio atlanticus LGP32 mutants lacking the genes involved in known quorum sensing pathways exhibit the same phenotype as wild-type Vibrio atlanticus LGP32, and Vibrio density does not induce attacks. However, V. atlanticus LGP32 produces AI-2 during attacks. Quorum sensing and iron acquisition are sometimes interconnected in Vibrio (McRose et al., 2018). For example, in Vibrio vulnificus, the production of vulnibactin (a siderophore) is known to be controlled by AI-2 (Kim and Shin, 2011). Similarly, AI-2 could be involved in the production of vibriferin in Vibrio atlanticus LGP32.

In the natural environment, associations between bacteria and algae have already been observed (Lopez-Joven et al., 2018; Miller et al., 2005; Rosales et al., 2022; Xu et al., 2022). We have shown here that algal attacks by vibrios can be carried out in vitro. Environnemental data collected in the Thau lagoon showed a correlation between the presence of Alexandrium and that of Vibrio, suggesting that such interactions could also occur in the marine environnement. If that were the case, this behaviour would provide an important ecological advantage to Vibrio to obtain nutrients in environment, where Alexandrium spp. and Gymnodinium catenatum form blooms.

With more than 30 species distributed all over the world (Anderson et al., 2012; Hallegraeff et al., 2021), Alexandrium spp. and Gymnodinium spp. considered as invasive species by the Delivering Alien Invasive Species Inventories for Europe (http://www.europe-aliens.org)), could play an unexpected and important role in maintaining, structuring and regulating Vibrio populations in the ecosystem. In turn, Vibrio could contribute to the regulation and control of their blooms.

To conclude, this study reveals the capacity of some Vibrio spp. to be facultative predators-like that hunt specific algae. In the current context of climate change favourable to their development monitoring the invasive algae, Alexandrium spp. and Gymnodinium catenatum should be considered not only for their potent harmful effect on humans and animals, but also as a potential source of nutrients for the expansion of Vibrio, particularly pathogenic species (Lemire et al., 2015).

eLife Assessment

This important study convincingly shows that Vibrio bacteria act as predators of ecologically significant algae that contribute to harmful blooms in the lab and in their natural habitat, and that predation is induced by starvation. The authors suggest a working model that can be the basis for future work on this system. The study will be very impactful to those interested in the diversity of microbial predator-prey interactions and controlling toxic algal bloom.

Reviewer #1 (Public review):

Summary:

Rolland and colleagues investigated the interaction between Vibrio bacteria and Alexandrium algae. The authors found a correlation between the abundance of the two in the Thau Lagoon and observed in the laboratory that Vibrio grows to higher numbers in the presence of the algae than in monoculture. Timelapse imaging of Alexandrium in coculture with Vibrio enabled the authors to observe Vibrio bacteria in proximity to the algae and subsequent algae death. The authors further determine the mechanism of the interaction between the two and point out similarities between the observed phenotypes and predator prey behaviours across organisms.

Strengths:

The study combines field work with mechanistic studies in the laboratory and uses a wide array of techniques ranging from co-cultivation experiments to genetic engineering, microscopy and proteomics. Further, the authors test multiple Vibrio and Alexandria species and claim a wide spread of the observed phenotypes.

Comments on revisions:

I thank the authors for their additional work on the manuscript. My comments were addressed to my satisfaction.

Reviewer #2 (Public review):

Goal summary

The authors sought to (i) demonstrate correlations between the dynamics of the dinoflagellate Alexandrium pacificum and the bacterim Vibrio atlanticus in natural populations, ii) demonstrate the occurrence of predation in laboratory experiments, iii) demonstrate that predation is induced by predator starvation, and iv) test for effects of quorum sensing and iron-uptake genes on the predation process.

Strengths include

- Data indicating correlated dynamics in a natural environment that increase the motivation for study of in vitro interactions

- Experimental design allowing clear inference of predation based on population counts of both prey and predators in addition to microscopy-based evidence

- Supplementation of population-level data with molecular approaches to test hypotheses regarding possible involvement of quorum sensing and iron update in predation

Weaknesses include

- A quantitative analysis of effects of manipulating V. atlanticus density on rates of predation would have been valuable

- Lack of clarity in some of the methodological descriptions

Appraisal

The authors convincingly demonstrate that V. atlanticus can prey on A. pacificum, provide strongly suggestive evidence that such predation is induced by starvation and clearly demonstrate that both iron availability and correspondingly the presence of genes involved in iron uptake strongly influence the efficacy of predation.

Discussion of impact

This paper will interest those interested in the diversity of forms of microbial predation and how microbial predatory behavior responds to environmental fluctuations. It will also interest those investigating bacteria-algae interactions and potential ecological controls of algal blooms. It may also interest researchers of microbial cooperation in light of the suggestion of communication between predator cells.

Author response:

The following is the authors’ response to the previous reviews

Public Reviews:

Reviewer #1 (Public review):

Summary:

Rolland and colleagues investigated the interaction between Vibrio bacteria and Alexandrium algae. The authors found a correlation between the abundance of the two in the Thau Lagoon and observed in the laboratory that Vibrio grows to higher numbers in the presence of the algae than in monoculture. Timelapse imaging of Alexandrium in coculture with Vibrio enabled the authors to observe Vibrio bacteria in proximity to the algae and subsequent algae death. The authors further determine the mechanism of the interaction between the two and point out similarities between the observed phenotypes and predator prey behaviours across organisms.

Strengths:

The study combines field work with mechanistic studies in the laboratory and uses a wide array of techniques ranging from co-cultivation experiments to genetic engineering, microscopy and proteomics. Further, the authors test multiple Vibrio and Alexandria species and claim a wide spread of the observed phenotypes.

Comments on revisions:

I thank the authors for their additional work on the manuscript. My comments were addressed to my satisfaction.

Dear Reviewer #1, we thank you for your careful evaluation of our manuscript and for the time and effort you dedicated to this review. We are pleased that the revised version has addressed your concerns to your satisfaction.

Reviewer #2 (Public review):

Goal summary

The authors sought to (i) demonstrate correlations between the dynamics of the dinoflagellate Alexandrium pacificum and the bacterim Vibrio atlanticus in natural populations, ii) demonstrate the occurrence of predation in laboratory experiments, iii) demonstrate that predation is induced by predator starvation, and iv) test for effects of quorum sensing and iron-uptake genes on the predation process.

Strengths include

- Data indicating correlated dynamics in a natural environment that increase the motivation for study of in vitro interactions

- Experimental design allowing clear inference of predation based on population counts of both prey and predators in addition to microscopy-based evidence

- Supplementation of population-level data with molecular approaches to test hypotheses regarding possible involvement of quorum sensing and iron update in predation

Weaknesses include

- A quantitative analysis of effects of manipulating V. atlanticus density on rates of predation would have been valuable

- Lack of clarity in some of the methodological descriptions

Appraisal

The authors convincingly demonstrate that V. atlanticus can prey on A. pacificum, provide strongly suggestive evidence that such predation is induced by starvation and clearly demonstrate that both iron availability and correspondingly the presence of genes involved in iron uptake strongly influence the efficacy of predation.

Discussion of impact

This paper will interest those interested in the diversity of forms of microbial predation and how microbial predatory behavior responds to environmental fluctuations. It will also interest those investigating bacteria-algae interactions and potential ecological controls of algal blooms. It may also interest researchers of microbial cooperation in light of the suggestion of communication between predator cells.

Dear Reviewer #2, we sincerely thank you for the time you devoted to this second review of our manuscript. We greatly appreciate your thoughtful comments, which helped us further improve the clarity and precision of the manuscript. All your additional recommendations have been carefully considered and addressed in the revised version and in our responses below.

Recommendations for the authors:

Reviewer #2 (Recommendations for the authors):

(2) The authors' reference to Fig. 4a did not address our concern about density potentially affecting the outcomes shown in Fig. 3. Fig. 4a does not provide any quantitative effects of manipulating Vibrio density. But the new density numbers the authors added in response to point (33) do seem to address our concern, because Vibrio densities become lower in the older cultures, excluding the possibility that the increased predation in older cultures might have been due higher Vibrio densities. We think this should be stated explicitly.

(33) See point (2) above. We think the authors should explicitly state in the text that the increased predation in older cultures was not due higher Vibrio densities in those older cultures, referring to their data.

As recommended by Reviewer#2, we added the sentence “Importantly, Vibrio densities decreased with culture age, ruling out the possibility that the stronger predation observed in older cultures was driven by higher bacterial densities” in the results section “Attack of A. pacificum ACT03 is activated by V. atlanticus LGP32 starvation.”

(45) Is it known that bacterial predators collectively feed more on other bacteria than on microbial eukaryotes in natural habitats? While this certainly seems most likely, it's stated as fact and so should either the statement should be supported with relevant citations or phrased as a likely hypothesis.

As suggested, we rephrased this sentence “Predatory bacteria are found in a wide variety of environments and are commonly described as feeding on other bacteria, although some cases of predation on microbial eukaryotes have also been hypothesized” in the discussion section.

(46) Perhaps "Conceiving predators as free-living organisms that kill other organisms and feed on them, this study suggest that Vibrios engage in a novel form of predation in which they kill and feed on algae."

The reference to 'developing' a predator behavior is not clear. What is meant by 'develop'? It seems unnecessary.

The use of italics when writing Vibrio is inconsistent.

We agree that the reference to “developing” a predatory behavior was unclear and unnecessary. We therefore revised the sentence as follows: “Conceiving predators as free-living organisms that kill other organisms and feed on them, this study suggests that Vibrio engages in a novel form of predation in which it kills and feeds on algae.” We also corrected the inconsistent use of italics for Vibrio throughout the manuscript.

(48) The authors might wish to revise this sentence, as although M. xanxthus does have contact-dependent killing mechanism, it is our understanding that both Lysobacter and myxobacteria can kill some prey at a distance with diffusible secretions.

The sentence “These bacteria must be in close proximity to their prey in order to cause lysis and utilize their biomass, regardless of the prey's species” was replaced by “These bacteria may require close proximity to their prey to cause lysis and utilize their biomass, although some can also kill prey at a distance through diffusible secretions”.

(50) Why not directly say 'predatory behavior?

We totally agree and have reworded the sentence.

Line by line feedback:

28 '...the phycosphere, an interface ...'

We agree and have revised the wording.

24 'In the attack stage, Vibrios...'

This sentence has been rephrased as recommended.

35 surrounds -> surround

The correction has been done.

36 The lysis is induced by the cells not by the 'stage'. We would rephrase to 'in which the lysis and consumption of the dinoflagellates occurs'

This sentence has been rephrased as recommended.

41 'a new mechanism that could to be involved' -> 'a new mechanism that could be involved ...'

The correction has been done.

61 forms

The correction has been done.

98 'the role...in'

The suggested correction has been performed.

103 'Qpcr' -> 'qPCR'

Thank you for spotting this typo. “Qpcr” was corrected to “qPCR” in the manuscript.

125 Misplaced punctuation

The punctuation was corrected.

152 The use of '.' vs 'x' to indicate multiplication when writing numbers is inconsistent. In some cases both are missing.

Numbers have been corrected throughout the manuscript.

231 I would rephrase 'poor nutrient stress' to 'little nutrient stress' or 'no nutrient stress'

The rephrasing was carried out as suggested.

310 R and used packages are not cited

We added the citation (R Core Team, 2024). Linear models, QQ plots (which are part of linear models), tests, and AICs are included in R by default and are credited to the R Core Team.

The sentence “Statistical analyses were performed using R 3.6.3 software” was replaced by “Statistical analyses were performed using R 3.6.3 software (R Core Team, 2024) using Rstudio”.

358 'are capable of simultaneously attacking'

The expression “are capable of simultaneously attacking” was revised in the manuscript to improve clarity and readability.

366 'exponential growth phase'

We have corrected the wording to “exponential growth phase” in the revised manuscript.

430 The large difference in incubation time between the sea-water vs nutrient-rich treatments and use of different media are unfortunate. These additional variables compromise the ability to directly ascribe observed differences to starvation.

We agree, the sentence “The comparative analysis of the proteome of V. atlanticus LGP32 incubated 60 h in artificial seawater (ENSW) versus V. atlanticus LGP32 grown 12 h in Zobell nutrient-rich medium revealed 10 proteins modulated by nutrient stress (Fig. S2)” was replaced by “The comparative analysis of the proteome of V. atlanticus LGP32 incubated 60 h in artificial seawater (ENSW) versus V. atlanticus LGP32 grown 12 h in Zobell nutrient-rich medium revealed 10 proteins that were differentially abundant under these two contrasting conditions (Fig. S2)”

443 Somewhat unclear sentence. I would rephrase this to "Remarkably, of the 10 proteins identified by proteomic analysis and eliminated by mutation, only elimination of PvuB prevented V. atlanticus from attacking A. pacificum ACT03."

To clarify this point, the sentence “Remarkably, among the 10 proteins identified by proteomic analysis only V. atlanticus LGP32 mutant lacking pvuB failed to attack A. pacificum ACT03 (Fig. 4C; ANOVA p <0.001)” was replaced by “Remarkably, of the 10 proteins identified by proteomic analysis and eliminated by mutation, only elimination of PvuB prevented V. atlanticus from attacking A. pacificum ACT03 (Fig. 4C; ANOVA p <0.001).”

445 'attack simultaneously' -> 'simultaneously attack'

The suggested modification has been done.

450 H3BO4 is written as Boron later, it would be good to call it boron here as well so that it is easier to make the connection for the reader.

We agree, we modified the manuscript and called it boron.

459 'no linked' -> 'no link'

The text was modified accordingly.

483 'which induces' -> 'which induce'

The correction has been made.

519 The use of Vibrio atlanticus and V. atlanticus is inconsistent within the text.

We have checked and modified the manuscript in accordance with the recommendations.

807-808 The use of the phrase 'Akaike information criterion (AICc) models' is confusing. Aren't these models just generalized linear models? It should be rephrased to make clear that the AICc is just a test that is used to select which model to use.

We clarified this point by revising Figure 1 legend. The sentences “(C) Result of Akaike information criterion (AICc) models tested to explain the mean value of degraded Alexandrium cells (dead cells) in spring. (D) Wald test of the AICc model attributing the mean value of degraded cells of Alexandrium in spring to free Vibrio “were replaced by “(C) Results of the Akaike Information Criterion (AICc) test conducted to select a model for explaining the mean value of dead Alexandrium (degraded cells) in spring. (D) Wald test of the AICc model explaining the mean value of dead Alexandrium in spring by free Vibrio”

827 The chronological sequence of snapshots is not very clear. Perhaps it would be clearer if pictures over a shorter timeframe were used to clearly show the gathering of the V. atlanticus cells near the algal cells.

To address this point, we removed the first and the last 14 seconds of the snapshots to clearly show the gathering of the V. atlanticus cells near the algal cells, and we added an arrow on Fig. 2D to indicate the chronological order.

Version 3

Abstract

Phytoplankton serve as a source of nutrients for bacteria in the marine environment. The interactions between algae and bacteria are known to include mutualism, commensalism, competition or antagonism. This occurs in the microenvironment surrounding phytoplankton cells, the phycosphere, an interface rich in nutrients and organic molecules exuded by the cell. Here, based on in situ observations and on an in vitro interaction study, we report on a novel form of starvation-induced hunting that the cells of selected Vibrio species exert on dinoflagellates. The results showed that Vibrio atlanticus was capable of attacking and killing the dinoflagellate Alexandrium pacificum ACT03. Briefly, the observed mechanism of algal-killing consists of first, the ‘immobilization stage’ involving the secretion of algicidal metabolites that disrupt the flagella of the algae. In the ‘attack stage’, Vibrios simultaneously surround algal cells at high density for a brief period without invading them. Finally, the ‘killing stage’ in which the lysis and consumption of the dinoflagellates occur. By using a combination of biochemical, proteomic, molecular and fluorescence microscopy approaches, we showed that this relationship is not related to the decomposition of algal organic matter, Vibrio quorum sensing pathways, toxicity of the algae or pathogenicity of the bacterium but is conditioned by nutrient stress, iron availability and linked to the iron-vibrioferrin transport system of Vibrio atlanticus. This is the first evidence of a new mechanism that could be involved in regulating Alexandrium spp. blooms and giving Vibrio a competitive advantage in obtaining nutrients from the environment. The interaction model we propose here suggests that Vibrio could play a role in regulating the proliferation of Alexandrium spp., giving it a competitive advantage in obtaining nutrients from the environment.

Introduction

Harmful algal blooms (HABs) have experienced an increase in their occurrence, intensity, and geographical distribution on a global scale, resulting in adverse environmental, health, and socioeconomic impacts (Marampouti et al., 2021). HABs have a considerable impact on human health as a result of direct exposure to volatile toxins or by toxic seafood consumption (Burkholder et al., 2018). From an ecological point of view, the expansion of HABs can result in the erosion of biodiversity, because they cause massive mortality of marine species and they are generally monospecific in nature (Chai et al., 2020). In coastal areas, understanding the biological interactions that control toxic algal blooms is therefore a major ecological challenge. Among HAB-causing organisms, a number of Alexandrium species have been placed on the list of invasive Mediterranean species. Among them, Alexandrium pacificum is a flagellated eukaryotic unicellular organism that together with Alexandrium tamarense and Alexandrium fundyense forms the "Alexandrium tamarense" complex responsible for paralytic shellfish poisoning worldwide (Hadjadji et al., 2020). Since 1998, A. pacificum (former A. catenella) was monitored by the French phytoplankton observation and monitoring network (Rephy) in the Thau lagoon (French Mediterranean) because it produces paralytic shellfish toxins (PSTs) resulting in paralytic shellfish poisoning (PSP) syndrome.

Laanaia (Laanaia et al., 2013) showed that in Thau lagoon, a water temperature around 20°C for several days and organic and inorganic nutrients in sufficient concentrations are parameters favoring the development of A. pacificum, whose massive blooms occur in autumn. Algal blooms are seasonal events resulting in a rapid increase in the concentration of a species of algae in an aquatic environment. Depending on the species of phytoplankton, tolerance to physicochemical parameters varies, which influences when blooms occur (Leblad et al., 2020). Interestingly, although the collapse of phytoplankton blooms has been previously attributed to viruses (Pal et al., 2020), some ecological studies have suggested an important role of algicidal bacteria (Su et al., 2007; Wang et al., 2010). Among them several are belonging to the Vibrio genus (Li et al., 2014; Wang et al., 2020).

Vibrio (class γ-proteobacteria) are common microorganisms in marine systems worldwide (Baker-Austin et al., 2017; Mavian et al., 2020), where they are important components of the food chain, particularly in biodegradation, nutrient regeneration and biogeochemical cycles (Oberbeckmann et al., 2012). Vibrio is one of the most studied bacterial taxa due to their ubiquity in coastal marine systems and their capacity to cause infections in humans and animals, leading sometimes to epizootic or zoonotic epidemics (LeRoux et al., 2015; Mavian et al., 2020). Vibrio are extremely adaptable to their environment (Johnson, 2013). The main factors influencing their occurrence and distribution in water are temperature, salinity, nutrient availability (Wang et al., 2020), multiple strategies such as biofilm formation on biotic and abiotic surfaces (Espinoza-Vergara et al., 2020), or interactions with a multitude of other organisms such as eukaryotic predators (Drebes Dörr and Blokesch, 2020) or plankton (Lopez-Joven et al., 2018) are used by Vibrio in the environment. There is also evidence that global climate change has increased Vibrio-associated illnesses affecting humans and animals (Brumfield et al., 2021; Muhling et al., 2017). However, the drivers and dynamics of Vibrio survival and propagation in the marine environment are not yet fully understood.

A substantial number of research articles have highlighted the potential of γ-proteobacteria to exert algicidal activity against dinoflagellates, supporting the hypothesis that γ-proteobacteria such as Vibrio play a role in the control of algal blooms in situ (Coyne et al., 2022). However, the mechanisms behind Vibrio-driven algal lysis in the environment remain to be elucidated. Particularly, it is unclear how in the water column, algicidal compounds secreted by bacteria can concentrate around the algae to exert their lytic effect.

This study aims to describe observations made in the natural environment between Vibrio bacteria and Alexandrium algal blooms, and to determine in vitro the main factors involved in this relationship. Using a combination of biochemical, proteomic, molecular, and fluorescence microscopy approaches, we explored the role in algal toxicity, bacterial pathogenicity and the quorum sensing pathway on this relationship and showed the important role of nutrient stress and the iron uptake pathway in this unique Vibrio/Alexandrium interaction.

Methods

Quantification of Alexandrium algae and Vibrio bacteria in the environment by qPCR

Seawater samples were collected in the Thau Lagoon (southern France, a shallow Mediterranean ecosystem open to the sea (Abadie et al., 1999), during spring and autumn 2015. Briefly, samples were collected from the subsurface (-50 cm) near an oyster table at a phytoplankton surveillance site (part of the REPHY network, N 43°26.058’ and E 003°39.878’). Once a week during spring and autumn 2015, during field sampling campaigns, 20 L of water was filtered on board through a 180 μm pore-size nylon membrane. At the laboratory, according to Lopez-Joven et al. (Lopez-Joven et al., 2018) seawater was fractionated into two size classes as follows: 2 L of the above filtrate was filtered through a 0.8 μm pore-size polycarbonate Whatman Nuclepore membrane to obtain organisms in the 0.8–180 μm range corresponding to plankton-associated Vibrio and living Alexandrium forms. Then, the filtrate from the 0.8 µm membrane was filtered again through a 0.2 μm pore-size polycarbonate Whatman Nuclepore membrane until the membrane was saturated. Alexandrium cells, ranging from 25 to 40 µm, belong to the microphytoplankton and are therefore retained in the 0.8–180 µm fraction. Any Vibrio cells potentially associated with or attached to Alexandrium cells will also be retained in this fraction. Vibrio cells are approximately 0.5-0.8 µm thick. The fraction between 0.2 and 0.8 µm therefore includes the free-living Vibrio. The bacterial population collected on 0.8-µm-pore-size filters was designated the particle-associated community, and the population on 0.2-µm-pore-size filters was designated as the free-living community. Membranes (in triplicate) were then conserved in 500 µL of 100% EtOH at -20°C. Environmental DNA (eDNA) was extracted from the MF Millipore membrane using the Macherey-Nagel NucleoSpin Tissue Kit and resuspended in 100 µL of water. The samples were then stored at -20°C after eDNA quantity and purity were assessed using a NanoDrop system (NanoDrop Technologies, Wilmington, DE, USA). PCR amplification reactions were done on a Roche LightCycler 480 Real-Time thermocycler (qPHD platform, University of Montpellier, France) using specific primer pairs (Table 3). Typically, the reactions contained 1 µL of template DNA (the DNA concentration for all samples varied from 1 to 40 μg mL−1), 0.5 µL of each primer (3.33 µM) and 4 µL of reaction mixture (SYBR Green Master Mix) in a total volume of 6 μL. The reaction parameters were as follows: 5 minutes at 95°C (initial denaturation) and 40 cycles of 10 s at 95°C (denaturation), 10 s at the corresponding hybridization temperature (Table 3) and 10 s at 72°C (elongation). Melting curve profiles were generated by increasing the temperature from 65°C to 95°C at 0.5°C s-1. Amplification products were analysed using LightCycler software (Roche Diagnostics). Vibrio spp. and A. pacificum and A. tamarense were quantified by constructing calibration curves based on DNA from the V. atlanticus LGP32 reference strain (former V. tasmaniensis LGP32) and from the A. pacificum reference strain (ACT03: A. catenella strain isolated from Thau in 2003) and the A. tamarense reference strain (ATT07: A. tamarense isolated from Thau in 2007) (not shown).

Strains and growth conditions

Vibrio strains. Wild-type and isogenic mutants of V. atlanticus LGP32 (Table 1) were used in this study. Deletion-mutants included ΔluxR, ΔluxS, ΔluxM and ΔpvuB isogenic strains. The ΔpvuB mutant was constructed here by allelic exchange as described previously by Le Roux (Le Roux et al., 2007). We also used V. atlanticus LGP32 carrying the pSW3654T-GFP plasmid (Le Roux et al., 2007), hereafter referred to as V. atlanticus LGP32-GFP. Bacterial strains were grown at 22 ± 1°C in Zobell medium (0.38 µM iron (III)). When needed, 25 μg mL-1 chloramphenicol (Cm) was added to cultures of V. atlanticus LGP32-GFP (Le Roux et al., 2007).

Phytoplankton strains. Non-axenic phytoplankton species (Table 2) were grown in batch culture in enriched natural sea water (ENSW, 6.55 µM iron (III)) with a salinity of 36 practical salinity units (PSU) at 22 ± 1°C under cool white fluorescent illumination (100 µmol photons m-2 s-1) and a 12 h:12 h light:dark cycle (Harrison et al., 1980). The algae were used for experiments in their exponential growth phase.

Co-culture assay

For each tested phytoplankton species (Table 2), 2 x 104 cells harvested in their exponential growth phase (doubling time between 5 and 7 days) were placed in 20 mL of ENSW medium in a 50 mL suspension culture flask (Cellstar® PS, Greiner bio-one). After incubation for 24 h at 22 ± 1°C under cool white fluorescent illumination, 40 µL of Vibrio strains (Table 1) grown for 12, 36, 60 and 156 h in Zobell medium or in Zobell medium supplemented with FeCl3 (6 µM iron (III)) or with boron (H3BO3; 0.47 mM) or the corresponding culture supernatant were added to the phytoplankton cells (Table 2). After incubation at 22 ± 1°C for 0, 15, 30, 45 and 60 min under cool white fluorescent lights, living, non-swimming, attacked and lysed phytoplankton cells were counted in a sedimentation chamber under an inverted microscope. The number of lysed cells corresponded to phytoplankton cells showing disrupted membranes. Non-swimming algae were not counted as lysed cells. For Vibrio analysis, 100 µL of a 1:10 serial dilution mixture in ENSW (from 10-2 to 10-10) was plated on Vibrio Selective TCBS (thiosulfate-citrate-bile salts-sucrose) agar (in triplicate). After incubation for 24 h at 22 ± 1°C, the number of living Vibrio cells was determined by counting colony-forming units (CFUs). The data come from three independent experiments using independent phytoplankton cultures and independent bacterial cultures.

Microscope observations

The dynamic of the interaction between A. pacificum ACT03 isolated from the French Thau Lagoon, south of France (Laabir et al., 2011) and V. atlanticus LGP32, which is an oyster pathogen isolated from the French Atlantic coast (Gay et al., 2004) and present in the Thau Lagoon (Lopez-Joven et al., 2018) was surveyed. As the A. pacificum ACT03 strain (Table 2) used in the study is not axenic, there is potential for bacteria other than V. atlanticus LGP32 to be present in the experiments. To elucidate the interaction without thoroughly accounting for the non-axenic cells, interaction was observed under a Zeiss Axio upright fluorescence microscope equipped with an AxioCamMRm 2 digital microscope camera using V. atlanticus LGP32 tagged with green fluorescent protein (GFP). Lasers were used at excitation wavelength (λex) 488 nm for GFP (emission wavelengths (λem): 505–530 nm) and λex 532 nm for plankton chloroplasts (λem 560–630 nm). Images were taken sequentially to avoid cross-contamination between fluorochromes. Sequences of images were merged during the Vibrio-Alexandrium interaction using ZEN 2012 (blue edition) software. Interaction events between Vibrio strains and phytoplankton strains were also observed under a Leica TCS SPE confocal laser scanning system connected to a Leica DM 2500 upright microscope camera (Montpellier RIO Imaging Platform, University of Montpellier, France).

Comparative proteomic analysis

Vibrio sampling and protein extraction

V. atlanticus LGP32 was grown for 60 h at 22°C in artificial seawater (high nutrient stress, cond. 1) or 12 h at 22°C in Zobell media (no nutrient stress, cond. 2). After 10 min centrifugation at 8000 rpm, crude protein extracts of V. atlanticus LGP32 in each culture condition (triplicates) were obtained by sonication on ice at 20% amplitude for 20 s in 200 µL of ice-cold denaturing buffer (7 M urea, 2 M thiourea, 4% CHAPS in 30 mM Tris-HCl, pH 8.5) and clarified by centrifugation at 2000 x g, 15 min, 4°C. The protein concentration of the supernatant was estimated using the 2D Quant Kit (Cytiva™, MERCK) and samples were stored at -80°C until use.

Two-dimensional gel electrophoresis (2D gel)

Protein extracts were individually analysed on 2D gel electrophoresis (6 gels per condition each corresponding to different biological replicates). To do so, 100 µg of proteins from each extract was added to rehydration buffer (7 M urea, 2 M thiourea, 4% CHAPS, 65 mM DTT) for a total volume of 350 µL. They were then individually loaded onto 17 cm isoelectric focusing strips (Bio-Rad) with a stabilized non-linear pH ranging from 3 to 10. Due to the high complexity of the protein profile in the acidic part (left) of the gel pH 3–10 (Fig. S3), we conducted additional ‘close-up’ analyses in gels using 17 cm isoelectric focusing strips (Bio-Rad) with a narrower, stabilized pH gradient ranging from 4 to 7. Strips were rehydrated passively for 5 h at 22°C, followed by active rehydration for 14 h under a 50 V current at 22°C (to help large proteins enter the strips). Thereafter, isoelectrofocusing was carried out using the following programme: 50 V for 1 h, 250 V for 1 h, 8000 V for 1 h and a final step at 8000 V for a total of 140 000 V.h with a slow ramping voltage (quadratic increasing voltage) at each step. Focused proteins were reduced by incubating the strip twice in equilibration buffer (1.5 M Tris, 6 M urea, 2% SDS, 30% glycerol; bromophenol blue, pH 8.8) containing DTT (130 mM) at 55°C. Then, they were alkylated by incubation with equilibration buffer containing iodoacetamide (135 mM) on a rocking agitator (400 rpm) at room temperature protected from light. Proteins were also separated according to their molecular weight (second dimension) on 12% acrylamide/0.32% piperazine diacrylamide gels run at 25 mA per gel for 30 min followed by 75 mA per gel for 8 h using a Protean II XL system (Bio-Rad). Gels were stained using an MS-compatible silver staining protocol and scanned using a ChemiDoc MP Imaging System (Bio-Rad) associated with Image Lab software version 4.0.1 (Bio-Rad).

Comparative bioinformatics analysis of 2D gels

Twelve gels (six per condition) were selected for comparative analysis on PD-Quest v. 7.4.0 (Bio-Rad) to identify changes in protein abundance between the proteomic profiles of V. atlanticus LGP32 cultured in contrasting nutrient conditions (ENSW/Zobell). Spots whose mean intensity across six replicates per strain was two times higher or lower than those from the other strain, with a P <0.01 (Mann-Whitney U-test), were considered significantly different in terms of abundance between the two conditions (quantitative difference). Differentially represented spots were then excised from the gels, destained, trypsin-digested and the obtained peptides were identified by tandem mass spectrometry (MS-MS) using the PISSARO platform facility (University of Rouen, France). To identify protein(s) present in each spot, the obtained peptides were compared with V. atlanticus LGP32 reference genome (https://vibrio.biocyc.org/). The genes whose peptides matched strongly were retrieved and used for a BLASTx query against non-redundant databases to determine the protein identity of the best match. A gene was considered as strongly matched when at least two peptides matched the sequence with a coverage of > 6%. Their theoretical isoelectric point (pI) and molecular weights were also calculated using the Expasy server (https://www.expasy.org/) to compare them with the location of the spot on the gel. Altogether, these complementary analyses made it possible to characterize the protein identity of each spot with confidence.

Gene expression analysis

Vibrio sampling and RNA extraction

V. atlanticus LGP32 was grown in Zobell media for 36 h and 60 h at 22°C (decline phase of growth, nutrient stress) or for 12 h at 22°C (exponential growth phase, no nutrient stress). Total RNA was isolated from V. atlanticus LGP32 using the standard TRIzol method (Invitrogen Life Technologies SAS, Saint-Aubin, France) and then treated with DNase (Invitrogen) to eliminate genomic DNA contamination. After sodium acetate precipitation, the quantity and quality of the total RNA were determined using a NanoDrop spectrophotometer and agarose gel electrophoresis. Following heat denaturing (70°C for 5 minutes), reverse transcription was performed using 1 μg of RNA prepared with 50 ng μL-1 oligo-(dT) 12–18 mer in a 20-μL reaction containing 1 mM dNTPs, 1 unit μL-1 RNAseOUT and 200 units μL-1 Moloney murine leukaemia virus reverse transcriptase (M-MLV RT) in reverse transcriptase buffer, according to the manufacturer’s instructions (Invitrogen Life Technologies SAS, Saint-Aubin, France).

PCR amplification

Amplification reactions were analysed using a Roche LightCycler 480 Real-Time thermocycler (Bio-Environnement platform, University of Perpignan, France). In this study, several PCR primer pairs were designed using Primer3 software (optimal primer size: 20 bases; Tm: 60°C; primer GC%: 50; 2GC clamp and product size range: 150–200 bp) and calibrated with V. atlanticus LGP32 genomic DNA (Table 1). To determine the qPCR efficiency of each primer pair used, standard curves were generated using seven serial dilutions of genomic DNA (10-2, 10-3, 10-4, 10-5, 10-7 and 10-8) (not shown); the qPCR efficiencies of the tested genes varied between 1.85 and 2.08 (Table 3). For gene expression, reverse transcription was performed with 1 µg of total RNA using random hexamers and SuperScript IV reverse transcriptase (Invitrogen). The total qPCR reaction volume was 10 μL and consisted of 5 μL of cDNA (diluted 1:5), 2.5 μL of SensiFAST SYBR No-ROX Mix (Bioline) and 100 nM or 300 nM PCR primer pair (Table 3). The reaction parameters were as follows: 2 min at 95°C (initial denaturation) and 40 cycles of 5 s at 95°C (denaturation), 10 s at 59°C (annealing) and 20 s at 72°C (elongation). The specificity of each PCR was checked by measuring fluorescent signals during melting curve analysis (PCR product heated from 65°C to 95°C continuously and slowly at 0.1°C s−1). Relative expression was calculated by normalization to the expression of two constitutively expressed housekeeping genes, namely, 6PKF (VS_2913) and CcmC (VS_0852), using the delta-delta threshold cycle (ΔΔCt) method (Pfaffl, 2001).

Detection of quorum-sensing signalling molecules

Vibrio culture

To detect the QS molecules (AI-2, AI-1 and CAI-1) V. atlanticus LGP32 was grown in Zobell media for 12 h (exponential growth phase, control) and 60 h (decline phase of growth, nutrient stress).

AI-2 analysis

Bioluminescence assay using the QS bioluminescent of Vibrio campbellii MM32 (luxN::Cm, luxS::Tn5Kan) was used to detect AI-2 molecules in culture supernatants. Briefly, Vibrio cultures were centrifuged at 17,000 x g for 10 min, and the resulting supernatants were filtered on 0.22 μm. Then 20 μL of the filtrates were mixed with 180 μL of V. campbellii MM32 diluted 1:5000 then incubated at 30°C and 100 rpm. Luminescence and cell density (OD620) were collected in triplicate and analysed according to Tourneroche et al.(Tourneroche et al., 2019).

AI-1 and CAI-1 extraction and LC-MS analysis

Chemical analyses were conducted with a Q Exactive Focus Orbitrap System coupled to an Ultimate 3000™ ultrahigh-performance liquid chromatography (UHPLC) system (Thermo Fisher Scientific) according to Rodrigues et al. (Rodrigues et al., 2022). Briefly, ethyl acetate (2 mL) was added into each culture (2 mL). This mixture was shaken overnight at room temperature (150 rpm). The two phases were then separated and the aqueous phase was extracted once again. The two obtained organic phases were pooled and the solvent was evaporated under vacuum. The crude extracts were dissolved in 500 µL LC-MS grade methanol for analysis. The experiments were performed using biological triplicates, each of which was analyzed in triplicate. Analyses of extracts and standards (3 μL injected) were performed in electrospray positive ionization mode in the 50–750 m/z range in centroid mode. The parameters were as follows: spray voltage: 3 kV; sheath flow rate: 75; aux gas pressure: 20; capillary temperature: 350°C; heater temperature: 430°C. The analysis was conducted in Full MS data-dependent MS2 mode (Discovery mode). Resolution was set to 70,000 in Full MS mode, and the AGC (automatic gain control) target was set to 1x106. In MS2, resolution was 17,500, AGC target was set to 2x105, isolation window was 0.4 m/z, and normalized collision energy was stepped to 15, 30 and 40 eV. The UHPLC column was a Phenomenex Luna Omega Polar C18 1.6 μm, 150 x 2.1 mm. The column temperature was set to 42°C, and the flow rate was 0.4 mL min-1. The solvent system was a mixture of water (A) with increasing proportions of acetonitrile (B), with both solvents modified with 0.1% formic acid. The gradient was as follows: 1% B 3 min before injection, then from 1 to 15 min, a gradient increase of B up to 100% (curve 5), followed by 100% B for 5 min. The flow was injected into the mass spectrometer starting immediately after injection. All data were acquired and processed using FreeStyle 1.5 software (Thermo Fisher Scientific).

Chemicals and solvents

N-acyl-homoserine lactones (AHL) were obtained from Cayman Chemical (Ann Arbor, MI, USA). Stock solutions were obtained by dissolving standards in methanol or dichloromethane (C18-AHL) at a concentration of 1 mg mL-1 and stored at -80°C. Standard solutions for UHPLC- high-resolution tandem mass spectrometry (HRMS) analyses were prepared by diluting each individual standard solution with methanol in order to obtain a concentration range from 2000 to 20 ng mL-1. LC-MS grade methanol, acetonitrile and formic acid were purchased from Biosolve (Biosolve Chimie, Dieuze, France), analytical-grade ethyl acetate was obtained from Sigma-Aldrich. Pure water was obtained from Elga Purelab Flex System (Veolia LabWater STI, Antony, France).

Nature of lytic compounds secreted by V. atlanticus LGP32

To determine the temperature-sensitivity of the lytic compounds secreted, V. atlanticus LGP32 grown for 60 h in Zobell media at 22 °C was filtered through a 10 kDa membrane (Amicon® Ultra-4 filter unit). The eluate containing molecules with MW below 10 kDa was then incubated in a water bath at 100°C for 30 min. Boiled filtrates (0.1 mL) were subsequently used to inoculate A. pacificum (ACT03 strain) cultures, then lytic activity was observed under the Leica TCS SPE confocal laser scanning system. Zobell media with the same treatment was used as control.

Statistical analysis

Environmental data

Statistical analyses were performed using R 3.6.3 software (R Core Team, 2024) using Rstudio. The relationship between Alexandrium and Vibrio was explored separately in spring and autumn. We used a generalized linear model specifying a Gaussian family. For spring, the dataset for salinity and temperature was complete (14 periods of observation). An influential period was detected and removed from the dataset because no dead Alexandrium cells were observed. The effects of explanatory variables such as log10 (Vibrio+1), salinity and temperature were centred, reduced and tested as fixed effects with a linear relationship. Model selection was performed using the Akaike information criterion corrected for small sample size (AICc). Models were considered different whenever the difference between their AICc value and the lowest AICc value (ΔAICc) was lower than 2 (Burnham and Anderson, 2002). Alexandrium distribution and model residuals were checked for normal distribution assumptions (QQ plot and Shapiro-Wilk test). For autumn, the dataset was complete for 10 periods of observation. Salinity and temperature were missing for three periods. We explored the relationship between Alexandrium and Vibrio alone using the method detailed above for spring.

In vitro data

Statistical analyses were performed using one-way ANOVA (analysed by pair) followed by Tukey’s test (Statistica 10.0 software, StatSoft, Maison-Alfort, France). *P <0.05, **P <0.01, ***P <0.001.

Results

Concomitant occurrence of A. pacificum ACT03, A. tamarense ATT07 and free-living Vibrio spp. in the Thau lagoon

In the spring and autumn of 2015, in the Thau Lagoon (Fig. 1A), we detected Alexandrium algae (A. pacificum ACT03 and A. tamarense ATT07, both alive and in degraded cell forms) and free-living Vibrio, but no plankton-associated Vibrio were observed (Fig. 1B, Data S1). Using model selection based on AICc, we found no significant relationship between Alexandrium (A. pacificum ACT03, A. tamarense) and Vibrio spp. abundances in autumn. This result is consistent with Vibrio’s difficulty in growing at temperatures below 20°C, as well as with the many environmental factors that can influence the dynamics of algae proliferation (Laanaia et al., 2013). Interestingly, in spring 2015, the mean densities of all Alexandrium cells (degraded and alive) and of free-living Vibrio were positively correlated. The lowest AICc was obtained with the model explaining degraded form of Alexandrium density based on the free Vibrio density (Fig. 1C). Given that, this model is not so different from the model with only the intercept, but better than any other linear combination with other potentially interfering drivers, such as temperature and salinity (Fig. 1D), we searched for evidence of a relationship between Vibrio and Alexandrium by studying their interaction in vitro.

Dynamics of Alexandrium and Vibrio in the environment.

(A) Location of the monitoring station in the Thau Lagoon (southern France). (B) Mean abundance (DNA equiv.) of Vibrio spp. (16S) and Alexandrium spp. (A. pacificum ACT03 + A. tamarense ATT07). Vibrio cells (black line with diamond dot) and degraded Alexandrium cells (grey line with round dot) were detected in the 0.2–0.8 μm fraction (free Vibrio fraction) in spring and autumn 2015. Living Alexandrium cells (grey line with round dot) but no plankton-associated Vibrio spp. (black line with diamond dot) were evidence in the 0.8–180 μm in spring and autumn. (C) Results of the Akaike Information Criterion (AICc) test conducted to select a model for explaining the mean value of dead Alexandrium (degraded cells) in spring. (D) Wald test of the AICc model explaining the mean value of dead Alexandrium in spring by free Vibrio.

V. atlanticus LGP32 feeds on Alexandrium pacificum ACT03

To investigate whether Alexandrium interacts with Vibrio, we incubated in Enriched Natural SeaWater (ENSW) A. pacificum ACT03 (2.0 x 104 cells) with V. atlanticus LGP32 previously grown for 12 hours in Zobell media (initial concentration of 8.8 x 107 cells mL-1). In interaction with V. atlanticus LGP32, A. pacificum ACT03 cell abundance decreased significantly from 2.1 x 104 cells mL-1 after 1h of exposure to 1.1 x 104 cells mL-1 after 48 h of exposure (Fig. 2A), while the V. atlanticus LGP32 concentration grew significantly after 26 h of interaction, reaching a maximum peak density of 7.6 x 107 CFU mL-1 at 34 h (Fig. 2B). In the control experiment where A. pacificum ACT03 was cultured alone in ENSW, the algal concentration remained stable over time (Fig. 2A) and no bacteria were on the corresponding TCBS plates (Vibrio selective medium). In the control where V. atlanticus LGP32 was grown alone in ENSW, the bacterial concentration decreased from 7.0 × 10L CFU/mL after 1 hour of incubation to 1.1 × 10L CFU/mL after 48 hours of incubation (Fig. 2B). These results show that the interaction between V. atlanticus LGP32 and A. pacificum ACT03 leads to a decline in the algal population and promotes the growth of V. atlanticus LGP32. This suggests that V. atlanticus LGP32 is able to feed on A. pacificum ACT03.

Incubation of V. atlanticus LGP32 and Alexandrium pacificum ACT03 in enriched natural seawater (ENSW).

(A) A. pacificum ACT03 cultured alone (grey bar) and incubated with V. atlanticus LGP32 (black bar) in ENSW. (B) V. atlanticus LGP32 cultured alone (grey bar) and incubated with A. pacificum ACT03 (black bar) in ENSW. (C) Snapshot of the interaction between V. atlanticus LGP32-GFP cells (60-hour culture) and one cell of A. pacificum ACT03 taken at 8h00 of co-culture. (D) Chronological snapshots of the interaction (70 pictures, one per second). V. atlanticus LGP32 (small green cells) and A. pacificum ACT03 cell (large red cell). All experiments were done in triplicate. Asterisks indicate significant differences in a multiple comparison test (One-way ANOVA with post hoc Tukey test), *P ≤ 0.05.

V. atlanticus LGP32 performs attacks on A. pacificum ACT03

Epifluorescence microscopy observation of GFP-labelled V. atlanticus LGP32 (previously grown in Zobell medium) in interaction showed that V. atlanticus LGP32 cells are capable of simultaneously attacking A. pacificum ACT03 cells (Fig. 2C and Video 1). The attacks were extremely rapid, with empty thecae (algal envelopes) observed in the medium after less than 60 s (Fig. 2D and Video 2). During the attack, V. atlanticus LGP32 did not invade the algal cell but remained clustered on the cell surface (Fig. 2C).

Attack of A. pacificum ACT03 is activated by V. atlanticus LGP32 starvation

In 2002, Martin hypothesized that nutritional stress induces bacteria to lyse algae (Martin, 2002). To test this hypothesis, we monitored V. atlanticus LGP32 behaviour in response to starvation (Fig. 3). We observed that V. atlanticus LGP32 in exponential growth phase (12 h of culture in Zobell medium, 8.8 x 107 cells mL-1) did not interact with A. pacificum ACT03 cells for the first hour of contact (Fig. 3A, B, C). In contrast, V. atlanticus LGP32 in the decline phase (36 h of culture in Zobell medium, 1.3 x 107 cells mL-1) induced a significant decrease in the number of motile algae cells by 8.9% after 15 min and by 43.3% after 60 min (Fig. 3A). This phenomenon corresponded to the degradation and/or disruption of algal flagella (Video 3). The flagella no longer functioned correctly, which caused irregular swimming of the algae (Video 3, left cell). This was followed by a complete cessation of swimming. When the flagellum detached from the algae (Video 3, right cell), the attack occurred. With starved V. atlanticus LGP32 (60 h of culture in Zobell medium, 0.6 x 107 cells mL-1), algae immobilization was fast and significant (91.4% in 15 min, Fig. 3A), and algae were attacked individually being targeted by V. atlanticus LGP32 cells (Video 4). The percentage of cells attacked and killed peaked at 30% after 15–30 min of contact (Fig. 3B and 3B1) and then decreased. After 1 h, attacks had stopped with approximately 40% of the algal cells still alive (Fig. 3B1 and 3C). Although it remains unclear whether the attacks occur during a specific phase of growth, it is evident that the cells are already weakened before attack as they have all lost their flagella. An old-starved culture of V. atlanticus LGP32 (126 h of culture in Zobell medium, < 0.1 x 107 cells mL-1) significantly immobilized A. pacificum ACT03 cells within a few minutes, with lysis occurring immediately (Fig. 3A and 3C), making it impossible to detect attacks by V. atlanticus LGP32 (Fig. 3B). The lysis phase corresponded to initial vesicle formation followed by the bursting of A. pacificum ACT03 cells (Fig. 3C and 3C1). Importantly, Vibrio densities decreased with culture age, ruling out the possibility that the stronger predation observed in older cultures was driven by higher bacterial densities.

Role of V. atlanticus LGP32 starvation in the interspecific interaction process.

Experiments were conducted by incubating A. pacificum ACT03 with V. atlanticus LGP32 previously grown for 12, 36, 60 and 126 h in Zobell medium. (A) Cumulative percentage of motile A. pacificum ACT03 cells. (B) Cumulative number of cells attacked by V. atlanticus LGP32 and (C) Cumulative cell lysis after 0, 15, 30, 45 and 60 minutes of interaction. Corresponding pictures showing (A1) Black arrows indicate unhooked and degraded flagellum from A. pacificum ACT03 flagellum, (B1) Chronological sequence of five snapshots showing V. atlanticus LGP32-GFP cells (60-hour culture) and A. pacificum ACT03 cells, during the first hour of their interaction. V. atlanticus LGP32 (small green cells), living A. pacificum ACT03 (large red cells) and dead A. pacificum ACT03 (large green cell). (C1) Black arrow 1 indicates vesicle formation on A. pacificum ACT03 cell and black arrow 2 indicates lysed A. pacificum ACT03 cell. All percentages were determined based on a minimum of 2,000 cells of A. pacificum ACT03. All experiments were done in triplicate. Asterisks indicate significant differences in a multiple comparison test (One-way ANOVA with post hoc Tukey test), ***P ≤ 0.001

We next tested whether this lytic effect was linked to Vibrio culture supernatant and mediated by thermostable molecule(s) secreted by Vibrio. The culture supernatant of starved culture of V. atlanticus LGP32 (36 h) filtered through a 10 kDa membrane and then incubated at 100°C for 30 min still possessed its lytic properties, indicating that the algicidal compounds produced by V. atlanticus LGP32 are small thermostable molecules unlikely to be lytic enzymes, or lysins able to digest the algae cell. However, these experimental observations clearly show the key role of nutrient limitation in triggering the attack behaviour and the secretion of lytic compounds of V. atlanticus LGP32.

Attack occurs on A. pacificum ACT03 in exponential phase of growth

Here, we wondered whether the live/dead status of algae is important for V. atlanticus LGP32-mediated attacks targeting. To this end, A. pacificum ACT03 in exponential growth phase was first exposed for 30 minutes to the supernatant of a 126-hour culture of V. atlanticus LGP32, which induced lysis of 70% of the A. pacificum ACT03 cells (Figures 3C and 3C1 (arrow 2) and Video 4). Next, cells of V. atlanticus LGP32 from a 60-hour culture, capable of attacking A. pacificum ACT03 cells (Fig. 3B), were added. For 1 hour of exposure, no attack was observed on the previously lysed algae. This result is similar to what is observed on the Video 1 with flash attacks only on immobilized, but not degraded A. pacificum ACT03 cells (red), and not on lysed cells (green). In addition, no attacks occurred on cells from an old A. pacificum ACT03 culture (1-month culture). Together, with the very short duration of attacks (Video 1), these results indicate that V. atlanticus LGP32 attacked exponentially growing cells of A. pacificum ACT03, but not decomposing algae, suggesting that this behaviour is not just an opportunistic response of heterotrophic bacteria to organic substrates.

Attack is independent of quorum sensing

Considering the simultaneous action of V. atlanticus LGP32 in attacks on A. pacificum ACT03, we tested whether the attack process depended on the key physiological mechanism that regulates many functions in marine microbial cells, quorum sensing (QS) (Lami, 2019; Papenfort and Bassler, 2016), a type of cell-cell communication. Although QS is a cell-density-dependent mechanism, our results showed no attack from a 12 h culture of V. atlanticus LGP32 up to a concentration of 4 x 106 Vibrio mL-1 (Fig 4A). Attacks were only observed with V. atlanticus LGP32 from a 60 h culture at low concentration of 5 x 103 Vibrio mL-1 to the highest concentration tested of 5 x 105 Vibrio mL-1 (Fig 4A), consistent with the hypothesis that the attacks were independent of Vibrio density.

Role of quorum sensing and the vibrioferrin iron uptake pathway in the interaction process.

(A) Effect of V. atlanticus LGP32 cell density on the attack process. A. pacificum ACT03 cells (1 x 103 cells mL-1) were incubated with V. atlanticus LGP32 grown for 60 hours in Zobell medium at concentrations ranging from 5.103 to 5 x 105 cells mL−1 (black bars). For comparison, A. pacificum ACT03 incubated with V. atlanticus LGP32 grown for 12 hours in Zobell medium at concentrations ranging from 5 x 103 to 4 x 106 cells mL−1 (grey bars). The image on the bars indicates either unaffected algae (live red algae) or algae attacked by Vibrio (algae covered with green vibrio cell) during the interaction (B) CqsS, luxM, luxN, luxS, and luxP quorum sensing and PvsA, PvuB and PvuA2 vibrioferrin pathway genes expression in V. atlanticus LGP32 grown for 12, 36 and 60 h in Zobell medium. (C) Effect of V. atlanticus LGP32 mutants on the attack process. Experiments were conducted by incubating A. pacificum ACT03 with V. atlanticus LGP32, V. atlanticus LGP32 tagged with GFP, V. atlanticus LGP32 washed with ENSW or V. atlanticus LGP32 mutant ΔPvuB, ΔluxM, ΔluxR and ΔluxS previously grown 60 h in Zobell media (control), The percentage of A. pacificum ACT03 attacked was determined during the first 30 min of exposure. (D) Effect of V. atlanticus LGP32 culture media composition on the attack process. Experiments were conducted by incubating A. pacificum ACT03 with V. atlanticus LGP32 grown 60 h in Zobell media supplemented with boron (H3BO3) or FeCl3. The results were compared with an exposure to V. atlanticus LGP32 grown 60 h in Zobell media. All percentages were determined based on a minimum of 2,000 cells of A. pacificum ACT03. All experiments were done in triplicate. Asterisks indicate significant differences in a multiple comparison test (One-way ANOVA with post hoc Tukey test), **P ≤ 0.01, ***P ≤ 0.001.

The analysis of the expression of genes involved in the known QS pathways in Vibrio cell (Fig. S3A), highlighted that only the AI-2 pathway was induced during nutrient stress of V. atlanticus LGP32, because only the expression of the AI synthase (LuxS) and its receptor (LuxP) increased significantly (Fig. 4B; ANOVA p <0.05). This was confirmed by a QS bioluminescence assay, which showed AI-2 molecules (unquantified) in the Zobell culture supernatant of V. atlanticus LGP32 after 60 h of culture but not after 12 h of culture and not in the ENSW supernatant of V. atlanticus LGP32 after 12 or 60 h of culture. UHPLC-HRMS/MS provided no evidence of detectable HAI-1 and CAI-1 in any experiments.

Targeted mutagenesis of key genes involved in two of the three known QS pathways in vibrios (Fig. S3), ΔluxM (HAI-1 production), ΔluxS (AI-2 production), and ΔluxR (main high-density QS regulator), did not result in any changes in the attack behavior of V. atlanticus LGP32 (Fig. 4C). Combined with the absence of overexpression of the CqsS gene (inducible by CAI-1) involved in the last known QS pathway in Vibrio (Fig. S3), these results indicated that the attack by V. atlanticus LGP32 is most likely unrelated to QS.

Attack related to the availability of iron

The comparative analysis of the proteome of V. atlanticus LGP32 incubated 60 h in artificial seawater (ENSW) versus V. atlanticus LGP32 grown 12 h in Zobell nutrient-rich medium revealed 10 proteins that were differentially abundant under these two contrasting conditions (Fig. S2). The two most down-regulated proteins correspond to be ß-ketoacyl-(acyl-carrier-protein) synthase II (-22-fold in ENSW compared to Zobell), a key regulator of bacterial fatty acid synthesis, and the dihydroorotase (-6.6-fold in ENSW compared to Zobell), an enzyme essential for pyrimidine biosynthesis and thus bacterial proliferation and growth. The low expression of these proteins in ENSW is consistent with V. atlanticus LGP32 nutritional starvation. The most up-regulated protein in starved V. atlanticus LGP32, with an increase of more than 6-fold, was glucosamine-6-phosphate deaminase, an enzyme involved in bacterial energy metabolism probably necessary for its survival. Among the other up-regulated proteins, one was an iron siderophore-binding protein (Spot 4413, Fig. S2A) corresponding to the vibrioferrin outer membrane receptor PvuB, whose gene is part of the pvu operons involved in iron transport (Fig. S3B). Interestingly, the corresponding gene pvuB as well as the vibrioferrin membrane receptor gene pvuA2 (Fig. S3B) were both significantly induced in Vibrio under nutrient stress (Fig. 4B; ANOVA p <0.01) but not the one involved in the vibrioferrin biosynthesis, pvsA (Fig. 4B). Remarkably, of the 10 proteins identified by proteomic analysis and eliminated by mutation, only elimination of PvuB prevented V. atlanticus from attacking A. pacificum ACT03 (Fig. 4C; ANOVA p <0.001). In the absence of the pvuB gene, V. atlanticus LGP32 was unable to simultaneously attack A. pacificum ACT03. In addition, V. atlanticus LGP32 cells that had been washed with ENSW to remove their culture supernatant metabolites also failed to attack A. pacificum ACT03 (Fig. 4C; ANOVA p <0.001), which is congruent with the hypothesis that attacks depend on the V. atlanticus LGP32 vibrioferrin transport system. Finally, attacks increased significantly when FeCl3 was added to the Vibrio culture medium (Fig. 4D) but not with boron known to be capable of being transported by vibrioferrin (see Fig. 4D). Taken together, those results are consistent with the hypothesis that attacks are regulated by iron.

Attack is a Vibrio spp. behaviour specific to Alexandrium spp

To evaluate the dinoflagellates specificity of the attack behaviour, a selection of Vibrio spp. was co-cultured with a selection of dinoflagellate strains commonly found in the Mediterranean Sea. The results showed that, among the Vibrio spp. tested (pathogenic or not) all, when under nutrient stress, were able to secrete algicidal compounds, immobilize, attack and lyse A. pacificum ACT03 cells (Table 1) and no link to their pathogenicity for fish of invertebrates was observed (Table 1). Among the sixteen dinoflagellates species tested, only Alexandrium spp. (non-toxic and paralytic shellfish toxin (PST) producers) and Gymnodinium catenatum (PST producer) were immobilized, attacked and lysed by V. atlanticus LGP32 (Table 2), but no link to PSTs was revealed (Table 2).

Discussion

Predation is a widespread mode of interaction for survival in the natural world (Finke and Denno, 2004; Sinclair et al., 2003). Predatory bacteria are found in a wide variety of environments and are commonly described as feeding on other bacteria, although some cases of predation on microbial eukaryotes have also been hypothesized (Johnke et al., 2014; Perez et al., 2016). Conceiving predators as free-living organisms that kill other organisms and feed on them, this study suggests that Vibrio engage in a novel form of predation in which they kill and feed on algae.

In fact, the strategy developed by Vibrio to kill algae may be reminiscent of strategies previously described in the prokaryotes (Johnke et al., 2014). As shown in Video 1, the interaction between V. atlanticus LGP32 and A. pacificum ACT03 proceeds in three stages (Fig. 5). The first stage, the ‘immobilization stage’, recalls the strategy used by Streptomyces to immobilize its prey (Kumbhar et al., 2014) based on the secretion of algicidal metabolites that disrupt the flagella. The second stage, the ‘attack stage’ corresponding to the physical contact between Vibrios and Alexandrium, is similar to the strategy used by Myxococcus xanthus and Lysobacter. These bacteria may require close proximity to their prey to cause lysis and utilize their biomass, although some can also kill prey at a distance through diffusible secretions ( Martin, 2002; Genovesi et al., 2013; Perez et al., 2016; Zhang et al., 2020). V. atlanticus LGP32 also surrounds A. pacificum ACT03 cells at high density for a very short time, but does not invade the algal cell. Visually, this phenomenon resembles bacteria clustering around lysed ciliate cells (Blackburn et al., 1998). The third stage, the ‘killing stage’, is similar to that of epibiotic bacterial predators, which induce the lysis of bacteria from the outside (Rashidan and Bird, 2001). Overall, these observations suggest that V. atlanticus LGP32 can exhibit a predatory behavior.

Schematic representation of a putative strategy developed by Vibrio spp. to feed on Alexandrium spp. and G. catenatum in the environment.

(A) Vibrio in the environment when subjected to starvation secrete non-protein lytic compounds. (B) Some of these lytic compounds degrade the flagella, immobilizing the alga (immobilization stage). (C) Then Vibrio swims and clusters around its prey (attack stage). (D) Lytic compounds released by Vibrio where able to concentrate around the algal cells, thereby lysing the algae (killing stage). (E) Feeding on the released nutrients, Vibrio multiply and then spread in the environment. Yellow clouds: Lytic compound release by Vibrio, Grey clouds: Algal nutrients released upon lysis.

The attack behaviour of V. atlanticus LGP32 was linked to iron absorption, as mutants with impaired iron absorption completely lost its ability to attack the algae. Iron is an essential element for growth in most organisms, including phytoplankton (Martin and Fitzwater, 1988) and bacteria (Neilands, 1981), and its concentration in seawater is known to be very low with measurements in the open ocean surface at 0.2 nM and in deep waters at 0.6 nM (Millero, 1998).

Moreover, its low solubility in seawater limits its availability (Bruland et al., 1994; Wu and Luther, 1994). To acquire iron, bacteria have developed systems based on the secretion (and subsequent uptake) of iron-chelating siderophores to obtain this element from the environment (Amin et al., 2009a). Therefore, many Vibrio spp. produce a siderophore known as vibrioferrin, which is synthesized and secreted by proteins encoded by the pvsABCDE gene cluster (Fig. S3). Given that boron is known for its role in regulating a global bacterial cellular response to phytoplankton and to bind to vibrioferrin (Romano et al., 2013; Weerasinghe et al., 2013), we tested its potential involvement in simultaneous Vibrio attacks. Compared to the Zobell control, no effect on the number of attacks was observed. For iron-vibrioferrin uptake, Vibrio parahaemolyticus uses a membrane siderophore receptor, called PvuA, which is coupled to an inner membrane ATP-binding cassette (ABC). This ABC transporter system, which is comprised of proteins encoded by the pvuABCDE gene cluster (Fig. S3), is required for the transport of the siderophore across the inner membrane (Tanabe et al., 2003). Siderophores are not only iron carriers but also important regulators of virulence (Miethke and Marahiel, 2007) and mediators of bacterial interaction with phytoplankton (Amin et al., 2009b; Kramer et al., 2020). We showed here a pivotal role of iron in the interaction between V. atlanticus LGP32 and A. pacificum ACT03. This mirrors the mutualistic interaction observed between Gymnodinium catenatum and Marinobacter (Amin et al., 2009b). In fact, in natural settings, the co-occurrence of Marinobacter and G. catenatum is suggested to depend on a mutually beneficial utilization of iron and carbon resources (Bolch et al., 2011). As in the present study, iron seems to play a key role in the interaction. Indeed, the labile iron released through the photolysis of ferric chelates with vibrioferrin providing a crucial iron source for phytoplankton, which need substantial amounts of iron to support carbon fixation through photosynthesis (Amin et al., 2009a; Yang et al., 2021). This fixed carbon, in turn, sustains the growth of both the phytoplankton and their associated bacterial counterparts (Amin et al., 2009b; Kramer et al., 2020). Interestingly, if a general nutrient deficiency causes attacks, iron supplementation increases this number of attacks (Figure 4D), suggesting the importance of iron absorption in the attack behavior. Future studies should determine whether a nutrient deficiency increases the iron absorption capacity of Vibrio bacteria and whether this could play a major role in the attack mechanism.

This study showed that quorum sensing is not involved in microbial attacks. Thus, V. atlanticus LGP32 mutants lacking the genes involved in known quorum sensing pathways exhibit the same phenotype as wild-type V. atlanticus LGP32, and Vibrio density does not induce attacks. However, V. atlanticus LGP32 produces AI-2 during attacks. Quorum sensing and iron acquisition are sometimes interconnected in Vibrio (McRose et al., 2018). For example, in Vibrio vulnificus, the production of vulnibactin (a siderophore) is known to be controlled by AI-2 (Kim and Shin, 2011). Similarly, AI-2 could be involved in the production of vibrioferrin in V. atlanticus LGP32.

In the natural environment, associations between bacteria and algae have already been observed (Lopez-Joven et al., 2018; Miller et al., 2005; Rosales et al., 2022; Xu et al., 2022). We have shown here that algal attacks by Vibrio can be carried out in vitro. Environmental data collected in the Thau lagoon showed a correlation between the presence of Alexandrium and that of Vibrio, suggesting that such interactions could also occur in the marine environment. If that were the case, this behaviour would provide an important ecological advantage to Vibrio to obtain nutrients in the environment, where Alexandrium spp. and Gymnodinium catenatum form blooms.

With more than 30 species distributed all over the world (Anderson et al., 2012; Hallegraeff et al., 2021), Alexandrium spp. and Gymnodinium spp. considered as invasive species by the Delivering Alien Invasive Species Inventories for Europe (http://www.europe-aliens.org)), could play an unexpected and important role in maintaining, structuring and regulating Vibrio populations in the ecosystem. In turn, Vibrio could contribute to the regulation and control of their blooms.

To conclude, this study reveals the capacity of some Vibrio spp. to act as facultative predator-like bacteria that hunt specific algae. In the current context of climate change, which is favourable to their development, monitoring the invasive algae Alexandrium spp. and Gymnodinium catenatum should be considered not only for their potent harmful effect on humans and animals, but also because they may represent a potential source of nutrients for the expansion of Vibrio, particularly pathogenic species (Lemire et al., 2015).